---
title: 'ncov-severity: all sequences'
author: "Cassia Wagner"
date: "5/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(bitmapType = 'cairo')
```

#### Load data
```{r}
df <- read_tsv(file = '../results/2020414_df.tsv')
```

#### Subsetting to Washington only sequences
```{r}
df_wa <- filter(df, division == 'Washington')
```

#### Missing values
```{r}
sum(is.na(df_wa))
sum(is.na(df_wa$clade))
df_wa$strain[is.na(df_wa$clade)] 
```
All of those sequences are listed in ncov/config/exclude.txt.
```{r}
df_wa=drop_na(df_wa)
```

#### Variable prep
```{r}
df_wa$clade[df_wa$clade =='S'] <- 'D' #renames clade to appropriate mutation

df_wa$age_bin <- factor(df_wa$age_bin, levels = c("10-20", "20-30", "30-40", "40-50", "50-60", "60-70", "70-80", "80-90", "over 90"), ordered = TRUE)
df_wa$clade <- factor(df_wa$clade, levels = c("D", "G"))

df_wa$age[df_wa$age_bin == '10-20'] <- 15
df_wa$age[df_wa$age_bin == '20-30'] <- 25
df_wa$age[df_wa$age_bin == '30-40'] <- 35
df_wa$age[df_wa$age_bin == '40-50'] <- 45
df_wa$age[df_wa$age_bin == '50-60'] <- 55
df_wa$age[df_wa$age_bin == '60-70'] <- 65
df_wa$age[df_wa$age_bin == '70-80'] <- 75
df_wa$age[df_wa$age_bin == '80-90'] <- 85
df_wa$age[df_wa$age_bin == 'over 90'] <- 91 #Better to be conservative
```


#### Data summary
```{r}
summary(df_wa)
```

#### How are the clades spread over time?
```{r}
hist_date = ggplot(df_wa, aes(x=date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Count") +
  xlab("Date")
hist_date
```

#### Is there a difference in Ct by clade?
```{r}
t.test(df_wa$avg_ct~df_wa$clade)
wilcox.test(formula = avg_ct ~ clade, data= df_wa, alternative = "two.sided")
```
```{r}
ct <- ggplot(df_wa, aes(x=clade, y=avg_ct), na.rm=TRUE) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(caption = "D: Mean Ct = 19.89,  G: Mean Ct = 18.57,  t-test p-value = 0.000009\nWilcoxon Rank Sum test p-value = 0.000008")
ct
```




#### Is there a difference in age by clade?
```{r}
t.test(df_wa$age~df_wa$clade)
wilcox.test(formula = age ~ clade, data= df_wa, alternative = "two.sided")
```
```{r}
age <- ggplot(df_wa, mapping = aes(x=clade, y=age_bin, color=clade, fill=clade)) + 
  geom_dotplot(binaxis="y", stackdir="center", binwidth=0.17, position = position_jitter(width = 0, height=0.3)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(caption = "D: Mean age = 57.41,  G: Mean age = 50.75,   t-test p-value = 0.0002,\nWilcoxon Rank Sum test p-value = 0.0003")
age
```


#### What if we restrict analyses by week?
```{r}
df_wa$week[df_wa$date < as.Date("2020-03-10")] = 1 
df_wa$week[df_wa$date < as.Date("2020-03-17") & df_wa$date > as.Date("2020-03-09")] = 2
df_wa$week[df_wa$date > as.Date("2020-03-16")] = 3
week_labels <- c("1" = "Mar 3-9", "2" = "Mar 10-16", "3" = "Mar 17-24")
```
```{r}
ct_by_week <- ggplot(df_wa, mapping = aes(x=clade, y=avg_ct)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  facet_wrap(~ week, labeller=labeller(week = week_labels))
ct_by_week
```
```{r}
age_by_week <- ggplot(df_wa, mapping = aes(x=clade, y=age_bin, color=clade, fill=clade)) + 
  geom_dotplot(binaxis="y", stackdir="center", binwidth=0.1, position = position_jitter(width =0, height = 0.3)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  facet_wrap(~ week, labeller=labeller(week = week_labels))
age_by_week
```

I'm concerned that early time points are biasing age of the D clade since initially D clade was more prevalent in Washington. Many early cases in Washington were at Life Care Center in Kirkland, so would increase average age of clade D. Also, at the beginning of March, Washington had just only realized that COVID-19 was circulating in the community, so initial cases sequenced might have been more severe than later cases and would also bias estimtes. Thus, I'm restricting the analysis from March 10 onward to reduce the time bias. At this time both clades are circulating in Washington.
```{r}
df_wa_later <- df_wa[df_wa$date > as.Date("2020-03-09"), ]
```

#### For data from March 10, is there difference in Ct by clade?
```{r}
t.test(df_wa_later$avg_ct~df_wa_later$clade)
wilcox.test(formula = avg_ct ~ clade, data= df_wa_later, alternative = "two.sided")
```
```{r}
ct_later <- ggplot(df_wa, aes(x=clade, y=avg_ct), na.rm=TRUE) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(caption = "D: Mean Ct = 19.57,  G: Mean Ct = 18.52,  t-test p-value = 0.0005\nWilcoxon Rank Sum test p-value = 0.0004") +
  labs(subtitle = "Sequences from March 10-24")
ct_later
```

#### For data from March 10 on, is there a difference in age by clade?
```{r}
t.test(df_wa_later$age~df_wa_later$clade)
wilcox.test(formula = age ~ clade, data= df_wa_later, alternative = "two.sided")
```
```{r}
age_later <- ggplot(df_wa, mapping = aes(x=clade, y=age_bin, color=clade, fill=clade)) + 
  geom_dotplot(binaxis="y", stackdir="center", binwidth=0.17, position = position_jitter(width = 0, height=0.3)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(caption = "D: Mean age = 54.57,  G: Mean age = 50.53,   t-test p-value = 0.006,\nWilcoxon Rank Sum test p-value = 0.009") +
  labs(subtitle = "Sequences from March 10-24")
age_later
```