---
title: "Analysis of D614G substitution on Ct and clinical outcomes in WA"
author: "Cassia Wagner"
date: "08/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev="CairoPNG")
options(bitmapType = 'cairo')
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gridExtra)
library(ggsignif)
library(ggpubr)
```
#### Load data:
```{r}
df <- read_tsv("../data/WA_df.tsv", col_types = list("c", "c", "c", "D", "c", "d", "c", "c", "c", "c", "c", "c", "D", "c", "d"))
```
```{r}
summary(df)
```

#### Data prep for analysis:
```{r}
df$primers <- factor(df$primers, levels = c("N1, N2", "ORF1ab", "Egene, RdRp", "cobas", "wadoh", "scan"))
df$source <- factor(df$source, levels = c("UW", "wadoh", "SFS"))
df$clade <- factor(df$clade, levels = c("D", "G"))
df$county <- factor(df$county)
df$sex <- factor(df$sex)
df$deceased <- factor(df$deceased, levels = c("yes", "no"))
df$site <- factor(df$site, levels = c("outpatient", "inpatient"))
df$critical_care <- factor(df$critical_care)
df$immunocompromised_cancer <- factor(df$immunocompromised_cancer)
df$age_bin = factor(df$age_bin, levels = c("under 10", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80-89", "90+"), ordered = TRUE)
df <- df %>%
  mutate(severe = ifelse(critical_care == "yes" | deceased == "yes", "yes", "no")) %>%
  mutate(severe = factor(severe)) %>%
  mutate(post_symp = date - symptom_onset) %>%
  mutate(decade_age = bin_age/10)

df$week[df$date < as.Date("2020-02-25")] = 1
df$week[df$date < as.Date("2020-03-03") & df$date > as.Date("2020-02-24")] = 2
df$week[df$date < as.Date("2020-03-10") & df$date > as.Date("2020-03-02")] = 3
df$week[df$date < as.Date("2020-03-17") & df$date > as.Date("2020-03-09")] = 4
df$week[df$date < as.Date("2020-03-24") & df$date > as.Date("2020-03-16")] = 5
df$week[df$date < as.Date("2020-03-31") & df$date > as.Date("2020-03-23")] = 6
df$week[df$date < as.Date("2020-04-07") & df$date > as.Date("2020-03-30")] = 7
df$week[df$date < as.Date("2020-04-14") & df$date > as.Date("2020-04-06")] = 8
df$week[df$date < as.Date("2020-04-21") & df$date > as.Date("2020-04-13")] = 9
df$week[df$date < as.Date("2020-04-28") & df$date > as.Date("2020-04-20")] = 10
df$week[df$date > as.Date("2020-04-27")] = 12

```
```{r}
summary(df)
```
```{r}
summary(df[df$source == "UW",])
```
```{r}
summary(df[df$source == "SFS",])
```
```{r}
summary(df[df$source == "wadoh",])
```

### CT, AGE, CLADE ANALYSIS
Note: I'm going to split analysis by SFS, UW, WA-DoH sequences because these are two totally sample sets, collected for different reasons and PCR amplifed in different labs. UW sequences are from samples ordered by a healthcare provider while SFS sequences are from people at-home who were (mostly) not symptomatic enough to seek clinical care. WA-DoH sequences are from a variety of sources -- hospitals, contact tracing, etc.


#### How are the clades spread over time?
```{r, echo=FALSE, fig.height=4, fig.width=8}
labels <- c(UW = "UW\nVirology", wadoh = "WA-DoH", SFS = "SCAN", "(all)" = "All")

clade_time <- df %>%
  ggplot(aes(x=date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.1, 0.9)) +
  ylab("Count") +
  xlab("Date") +
  facet_grid(source ~., margins=TRUE, labeller = labeller(source=labels))
clade_time
```

#### How are UW Virology primers spread over time?
```{r, echo=FALSE, fig.height=4, fig.width=8}
primer_time <- df %>%
  filter(source == "UW") %>%
  ggplot(aes(x=date, color=primers, fill=primers)) + 
  geom_bar(position = position_stack()) + 
  theme_minimal() +
  theme(legend.position = c(0.85, 0.85)) +
  ylab("Count") +
  xlab("Date")
primer_time
```
Do different primers have different Ct's?
```{r, fig.height=4, fig.width=8}
dotplot_ct_primer <- df %>%
  filter(source == "UW") %>%
  ggplot(mapping = aes(x=primers, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, aes(color=primers, fill=primers)) +
  theme_minimal() +
  stat_summary(fun.data="mean_sdl", geom="pointrange", size = 0.1) +
  ylab("Average Ct") +
  xlab("Primer set") +
  geom_signif(comparisons = list(c("Egene, RdRp", "N1, N2"), c("N1, N2", "ORF1ab")), textsize = 3, test = "wilcox.test")
dotplot_ct_primer
```

For UW Virology samples, decided to focus on N1, N2 primers since they're the most abundant and the main primers used when D614 & G614 virus are circulating. However, I'll also include analysis with ORF1ab primers. There are not enough Egene, RdRp primers to do analysis with.

#### Is there a difference in Ct by clade?
For UW:
```{r}
t.test(df$avg_ct[df$source == "UW" & df$primers == "N1, N2"]~df$clade[df$source == "UW" & df$primers == "N1, N2"])
wilcox.test(formula = avg_ct ~ clade, data= df[df$source == "UW" & df$primers == "N1, N2", ], alternative = "two.sided")

medD_n1n2 = median(df$avg_ct[df$source == "UW" & df$primers == "N1, N2" & df$clade == "D"])
medG_n1n2 = median(df$avg_ct[df$source == "UW" & df$primers == "N1, N2" & df$clade == "G"])
medD_n1n2
medG_n1n2
medD_n1n2 - medG_n1n2
```

```{r}
t.test(df$avg_ct[df$source == "UW" & df$primers == "ORF1ab"]~df$clade[df$source == "UW" & df$primers == "ORF1ab"])
wilcox.test(formula = avg_ct ~ clade, data= df[df$source == "UW" & df$primers == "ORF1ab", ], alternative = "two.sided")

medD_orf1 = median(df$avg_ct[df$source == "UW" & df$primers == "ORF1ab" & df$clade == "D"])
medD_orf1
medG_orf1 = median(df$avg_ct[df$source == "UW" & df$primers == "ORF1ab" & df$clade == "G"])
medG_orf1
medD_orf1 - medG_orf1
```

For SFS:
```{r}
t.test(df$avg_ct[df$source == "SFS"]~df$clade[df$source == "SFS"])
wilcox.test(formula = avg_ct ~ clade, data= df[df$source == "SFS", ], alternative = "two.sided")

medD_sfs = median(df$avg_ct[df$source == "SFS" & df$clade == "D"])
medG_sfs = median(df$avg_ct[df$source == "SFS" & df$clade == "G"])
medD_sfs
medG_sfs
medD_sfs -medG_sfs
```

For WA-DoH:
```{r}
t.test(df$avg_ct[df$source == "wadoh"]~df$clade[df$source == "wadoh"])
wilcox.test(formula = avg_ct ~ clade, data= df[df$source == "wadoh", ], alternative = "two.sided")

medD_wadoh = median(df$avg_ct[df$source == "wadoh" & df$clade == "D"])
medG_wadoh = median(df$avg_ct[df$source == "wadoh" & df$clade == "G"])
medD_wadoh
medG_wadoh
medD_wadoh - medG_wadoh
```

```{r, fig.width=5}
dotplot_ct <- df %>%
  filter(primers == "N1, N2" | source == "SFS" | source == "wadoh") %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, binwidth = 0.5, stackratio=0.75, aes(color=clade, fill=clade)) +
  geom_boxplot(alpha=0, size=0.3) +
  scale_fill_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank(),
        panel.spacing = unit(-0.5, "lines")) +
  ylab("Average Ct") +
  xlab("") +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="wilcox.test") +
  facet_wrap(~ source, labeller = labeller(source=labels))
dotplot_ct
```

```{r, fig.width=5}
primer_ct <- df %>%
  filter(source == "UW") %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, binwidth = 0.5, aes(color=clade, fill=clade)) +
  geom_boxplot(alpha=0, size=0.3) +
  scale_fill_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank(),
        panel.spacing = unit(-0.5, "lines")) +
  ylab("Average Ct") +
  xlab("") +
  labs(subtitle = "All UW Virology samples") +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="wilcox.test") +
  facet_wrap(~ primers, labeller = labeller(source=labels))
primer_ct
```
```{r, fig.width=5}
orf1_ct <- df %>%
  filter(primers == "ORF1ab") %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, binwidth = 0.5, aes(color=clade, fill=clade)) +
  geom_boxplot(alpha=0, size=0.3) +
  scale_fill_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank(),
        panel.spacing = unit(-0.5, "lines")) +
  ylab("Average Ct") +
  xlab("") +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="wilcox.test") 
orf1_ct
```


#### Is there a difference in age by clade?
For UW:
```{r}
t.test(df$bin_age[df$source == 'UW']~df$clade[df$source == 'UW'])
wilcox.test(formula = bin_age ~ clade, data= df[df$source == 'UW', ], alternative = "two.sided")
```
For SFS:
```{r}
t.test(df$bin_age[df$source == 'SFS'] ~ df$clade[df$source == 'SFS'])
wilcox.test(formula = bin_age ~ clade, data= df[df$source == 'SFS', ], alternative = "two.sided")
```

Do we have the power to find an age difference in SCAN samples?
```{r}
testing_age <- function(df, ss) {
  ttest <- df %>%
    sample_n(ss) %>%
    t.test(bin_age ~ clade, data = ., var.equal = TRUE)
  return(ttest$p.value)
}

power_age_79 <- map(1:10000, ~ testing_age(df[df$source == "UW", ], 79))
```
```{r}
sum(power_age_79 < 0.05)/10000
```

```{r, fig.width=5}
dotplot_age <- df %>%
  drop_na(bin_age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=bin_age), na.rm=TRUE) + 
  geom_dotplot(aes( color=clade, fill=clade), binaxis="y", stackdir="center", stackratio = 0.3, dotsize = 0.5, position = position_jitter(width = 0, height=4)) +
  scale_fill_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank()) +
  ylab("Age") +
  xlab("") +
  stat_summary(fun.data="mean_sdl", geom="pointrange", size = 0.1) +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="t.test") +
   facet_wrap(~ source, labeller = labeller(source=labels))
dotplot_age
```
Note: We not have age for WA-DOH samples.


#### Age vs. Ct
```{r, fig.width=7}
ct_age <- df%>%
  filter(primers == "N1, N2" & !is.na(bin_age)) %>%
  ggplot(aes(x = bin_age, y = avg_ct)) +
  geom_point(aes(color=clade, fill=clade)) +
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  geom_smooth(colour = "black", size=0.5, method = "lm") +
  theme_minimal() +
  xlab("Age") +
  ylab("Average Ct")
ct_age
```

#### Age vs. time
```{r, fig.width=7}
age_time <- df %>%
  drop_na(age_bin) %>%
  ggplot(aes(x = date, y = bin_age)) +
  geom_point(aes(color=clade, fill=clade)) +
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  geom_smooth(method="lm", col = "black", size=0.5) +
  xlab("Date") +
  ylab("Age")
age_time

```


#### Ct vs. time
```{r, fig.width=7}
ct_time <- df %>%
  filter(primers == "N1, N2") %>%
  ggplot(aes(x = date, y = avg_ct)) +
  geom_point(aes(color=clade, fill=clade)) +
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  geom_smooth(method="lm", col = "black", size=0.5) +
  xlab("Date") +
  ylab("Average Ct")
ct_time
```

#### Ct vs. symptom onset
```{r, fig.width=7}
ct_sympon <- df %>%
  filter(!is.na(post_symp)) %>%
  ggplot(aes(x = post_symp, y = avg_ct)) +
  geom_point(aes(color=clade, fill=clade)) +
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  geom_smooth(data = subset(df, post_symp >=0 & post_symp < 20),method = "lm", color = "black", size=0.5) +
  xlab("Days since symptoms started") +
  ylab("Average Ct")
ct_sympon
```
```{r}
  cor.test(as.numeric(df$post_symp[df$post_symp >= 0 & df$post_symp < 20]), df$avg_ct[df$post_symp >= 0 & df$post_symp < 20], method = "pearson")

```


#### Linear model of Ct w/ UW
```{r}
lreg_ct <- glm(avg_ct ~ clade + bin_age + primers + week, data = df[df$source == "UW", ])
summary(lreg_ct)

df$residuals_ct[!is.na(df$age_bin) & df$source == "UW"] = residuals(lreg_ct)

p1_ct <- cbind(Fitted = fitted(lreg_ct), Residuals=residuals(lreg_ct)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_ct <- df %>%
  filter(!is.na(residuals_ct)) %>%
  ggplot(aes(x=residuals_ct)) +
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  xlab("Residuals") +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$residuals_ct[!is.na(df$residuals_ct)])))

gridExtra::grid.arrange(p1_ct, p2_ct, nrow=1)
```
It's probably not appropriate to just include primer type as separate predictors because these primers could have different intercepts and slopes. Therefore, I'll do it for N1, N2 primers and ORF1ab below.

```{r}
lreg_ct_n1n2 <- glm(avg_ct ~ clade + bin_age + week, data = df[df$source == "UW" & df$primers == "N1, N2", ])
summary(lreg_ct_n1n2)

df$residuals_ct_n1n2[!is.na(df$age_bin) & df$source == "UW" & df$primers == "N1, N2"] = residuals(lreg_ct_n1n2)

p1_ct_n1n2 <- cbind(Fitted = fitted(lreg_ct_n1n2), Residuals=residuals(lreg_ct_n1n2)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_ct_n1n2 <- df %>%
  filter(!is.na(residuals_ct_n1n2)) %>%
  ggplot(aes(x=residuals_ct_n1n2)) +
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$residuals_ct_n1n2[!is.na(df$residuals_ct_n1n2)])))

gridExtra::grid.arrange(p1_ct_n1n2, p2_ct_n1n2, nrow=1)

```

##### What would this change in Ct correspond to in viral load? 
Pavitra's standard curve for N1, N2 does not include any samples with Ct less than 20 and our average Ct is lower than that. So the curve may not be linear with lower Ct. But if we were to assume the curve was linear with lower Ct, the slope for {log(viral load) = Ct(slope) + intercept} is -0.2995.
```{r}
n1n2_slope = -0.2995
log_viral_load = lreg_ct_n1n2$coefficients["cladeG"]*n1n2_slope
log_viral_load
se = summary(lreg_ct_n1n2)$coefficients[2, 2]

upper_ci = (lreg_ct_n1n2$coefficients["cladeG"] + (1.96*se)) * n1n2_slope
lower_ci = (lreg_ct_n1n2$coefficients["cladeG"] - (1.96*se)) * n1n2_slope
upper_ci
lower_ci
```


```{r}
lreg_summary <- tibble("Variable" = c("Intercept", "614G", "Age", "Week"), 
                       "Coeff. est." = c("20.14", "-1.39", "0.0088", "-0.20"), 
                       "Std. Error" = c("0.66", "0.26", "0.0058", "0.12"),
                       "p-value" = c("<2e-16***", "9.7e-08***", "0.13", "0.092"))


table <- ggtexttable(lreg_summary, rows = NULL, 
                     theme = ttheme("lBlack", padding = unit(c(4, 3), "mm"), colnames.style = colnames_style(fill = "white", hjust=0, x =0.05, size=10), 
                                    tbody.style = tbody_style(fill = "white", hjust=0, x =0.05, size=10))) 

table
```


```{r}
lreg_ct_orf1 <- glm(avg_ct ~ clade + bin_age + week, data = df[df$source == "UW" & df$primers == "ORF1ab", ])
summary(lreg_ct_orf1)

df$residuals_ct_orf1[!is.na(df$age_bin) & df$source == "UW" & df$primers == "ORF1ab"] = residuals(lreg_ct_orf1)

p1_ct_orf1 <- cbind(Fitted = fitted(lreg_ct_orf1), Residuals=residuals(lreg_ct_orf1)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_ct_orf1 <- df %>%
  filter(!is.na(residuals_ct_orf1)) %>%
  ggplot(aes(x=residuals_ct_orf1)) +
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$residuals_ct_orf1[!is.na(df$residuals_ct_orf1)])))

gridExtra::grid.arrange(p1_ct_orf1, p2_ct_orf1, nrow=1)
```
Residual distribution does not look good here, which suggets that this GLM model is not appropriate with these primers.


```{r}
lreg_summary_orf1 <- tibble("Variable" = c("Intercept", "614G", "Age", "Week"), 
                       "Coeff. est." = c("23.81", "-2.64", "0.031", "-0.39"), 
                       "Std. Error" = c("2.71", "0.86", "0.018", "0.34"),
                       "p-value" = c("3.8e-16***", "0.0025**", "0.092", "0.26"))


table_orf1 <- ggtexttable(lreg_summary_orf1, rows = NULL, 
                     theme = ttheme("lBlack", padding = unit(c(4, 3), "mm"), colnames.style = colnames_style(fill = "white", hjust=0, x =0.05, size=10), 
                                    tbody.style = tbody_style(fill = "white", hjust=0, x =0.05, size=10))) 

table_orf1
```


### CLINICAL ANALYSIS
```{r}
summary(df[!is.na(df$deceased), ])
```

#### How are the clades spread over time for clinical samples?
```{r, echo=FALSE}
clade_time_uwmed <- df %>%
  filter(!is.na(deceased)) %>%
  ggplot(aes(x=date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  ylab("Count") +
  xlab("Date") +
  labs(subtitle = "Samples from patients in UW Medicine System") +
  facet_grid(primers ~ ., margins = TRUE)
clade_time_uwmed
```

#### Is Ct of UW Med affiliated patients different than Ct of other UW Virology patients?
```{r}
df_uw <- df%>%
  filter(source == "UW") %>%
  mutate(uw = ifelse(!is.na(deceased), "UW affiliated", "Other healthcare network" ))
```
```{r}
df_uw %>%
  filter(primers == "N1, N2") %>%
  {t.test(.$avg_ct ~ .$uw)}
wilcox.test(avg_ct ~ uw, data = df_uw[df_uw$primers == "N1, N2", ])

median(df_uw$avg_ct[df_uw$uw == "Other healthcare network" & df_uw$primers == "N1, N2"])
median(df_uw$avg_ct[df_uw$uw == "UW affiliated" & df_uw$primers == "N1, N2"])
```
```{r}
df_uw %>%
  filter(uw == "Other healthcare network" & primers == "N1, N2") %>%
  wilcox.test(avg_ct ~ clade, data = .)
```
```{r}
dotplot_ct_system <- df_uw %>%
  filter(primers == "N1, N2") %>%
  ggplot(mapping = aes(x=uw, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, binwidth = 0.5, fill="white") +
  geom_boxplot(alpha=0, size=0.3) +
  theme_minimal() +
  ylab("Average Ct") +
  xlab("") +
  geom_signif(comparisons = list(c("Other healthcare network", "UW affiliated")), textsize = 3, test="wilcox.test") 
dotplot_ct_system
```

#### Is Ct different across clade in UW Med affiliated patients?
```{r}
df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  {t.test(.$avg_ct ~ .$clade)}

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
wilcox.test(formula = avg_ct ~ clade, data = ., alternative = "two.sided")

median(df$avg_ct[!is.na(df$deceased) & df$primers == "N1, N2" & df$clade == "D"])
median(df$avg_ct[!is.na(df$deceased) & df$primers == "N1, N2" & df$clade == "G"])
```
```{r}
dotplot_ct_uwmed <- df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 1, binwidth = 0.5, aes(color=clade, fill=clade)) +
  geom_boxplot(alpha=0, size=0.3) +
  scale_fill_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade",values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank()) +
  ylab("Average Ct") +
  xlab("") +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="wilcox.test") +
  labs(subtitle = "Patients at UW Medicine affiliates")
dotplot_ct_uwmed
```


```{r, fig.width=5}
primer_ct_clinic <- df %>%
  filter(source == "UW" & !is.na(deceased)) %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, binwidth = 0.5, aes(color=clade, fill=clade)) +
  geom_boxplot(alpha=0, size=0.3) +
  scale_fill_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank(),
        panel.spacing = unit(-0.5, "lines")) +
  ylab("Average Ct") +
  xlab("") +
  labs(subtitle = "Samples from Patients in UW Medicine network") +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="wilcox.test") +
  facet_wrap(~ primers, labeller = labeller(source=labels))
primer_ct_clinic
```

#### Is their a difference in age between UW Medicine patients and other patients tested at UW Virology?
```{r}
t.test(df_uw$bin_age ~ df_uw$uw)
```

#### Is their a difference in age by clade in UW Medicine affiliated patients?
```{r}
df %>%
  filter(!is.na(deceased)) %>%
  {t.test(.$bin_age ~ .$clade)}
```
```{r, fig.width=4, fig.height=4}
dotplot_age_uwmed <- df %>%
  filter(!is.na(deceased)) %>%
  drop_na(bin_age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=bin_age), na.rm=TRUE) + 
  geom_dotplot(aes( color=clade, fill=clade), binaxis="y", stackdir="center", stackratio = 0.8, dotsize =0.8, position = position_jitter(width = 0, height=4)) + 
  scale_fill_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank()) +
  ylab("Age") +
  xlab("") +
  stat_summary(fun.data="mean_sdl", geom="pointrange", size = 0.1) +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="t.test") +
  labs(subtitle = "Patients at UW Medicine affiliates")
dotplot_age_uwmed
```

#### Across what variables in UW Medicine affiliated patients does Ct differ?
```{r, fig.width=5}
df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = county, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")

df%>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = bin_age, y = avg_ct)) +
  geom_point(aes(color=clade, fill=clade)) +
  geom_smooth(colour = "black", size=0.5, method = "lm")

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = sex, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = immunocompromised_cancer, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = site, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = severe, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")
```

#### Linear model of Ct in patients affiliated with UW Medicine
```{r}
lreg_ct_uwmed_n1n2 = glm(avg_ct ~ clade + sex + bin_age + week + immunocompromised_cancer + site + severe, data = df[!is.na(df$deceased) & df$primers == "N1, N2", ])

summary(lreg_ct_uwmed_n1n2)

df$residuals_ct_uwmed_n1n2[!is.na(df$bin_age) & !is.na(df$deceased) & df$primers == "N1, N2"] = residuals(lreg_ct_uwmed_n1n2)

p1_ct_uwmed_n1n2 <- cbind(Fitted = fitted(lreg_ct_uwmed_n1n2),Residuals=residuals(lreg_ct_uwmed_n1n2)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_ct_uwmed_n1n2 <- df %>%
  filter(!is.na(residuals_ct_uwmed_n1n2)) %>%
  ggplot(aes(x=residuals_ct_uwmed_n1n2)) +
  geom_histogram(aes(y = ..density..), binwidth = 1.5) +
  xlab("Residuals") +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$residuals_ct_uwmed_n1n2[!is.na(df$residuals_ct_uwmed_n1n2)])))

gridExtra::grid.arrange(p1_ct_uwmed_n1n2, p2_ct_uwmed_n1n2, ncol=2)
```
```{r}
lreg_summary_clinical <- tibble("Variable" = c("Intercept", "614G","Male", "Age", "Week", "Immunocomp. or\nactive cancer", "Hospitalized", "Critical care\nor deceased"), 
                       "Coeff. est." = c("16.00", "-1.35", "1.09", "0.012", "0.32", "-0.25", "1.22", "-0.65"), 
                       "Std. Error" = c("1.45", "0.55", "0.48", "0.014", "0.27", "0.77", "0.77", "0.98"),
                       "p-value" = c("<2e-16***", "0.014*", "0.024*", "0.36", "0.23", "0.74", "0.11", "0.51"))


table_clinical <- ggtexttable(lreg_summary_clinical, rows = NULL, 
                     theme = ttheme("lBlack", padding = unit(c(4, 3), "mm"), colnames.style = colnames_style(fill = "white", hjust=0, x =0.05, size=10), 
                                    tbody.style = tbody_style(fill = "white", hjust=0, x =0.05, size=10))) 

table_clinical
```


```{r}
lreg_ct_uwmed = glm(avg_ct ~ clade + sex + week + severe + bin_age + site + immunocompromised_cancer, data = df[!is.na(df$deceased) & df$primers == "ORF1ab", ])

#lreg_ct_uwmed = glm(avg_ct ~ clade + sex + week + severe + primers + site, data = df[!is.na(df$deceased), ])

summary(lreg_ct_uwmed)

df$residuals_ct_uwmed[!is.na(df$bin_age) & !is.na(df$deceased) & df$primers == "ORF1ab"] = residuals(lreg_ct_uwmed)

#df$residuals_ct_uwmed[!is.na(df$bin_age) & !is.na(df$deceased)] = residuals(lreg_ct_uwmed)

p1_ct_uwmed <- cbind(Fitted = fitted(lreg_ct_uwmed),Residuals=residuals(lreg_ct_uwmed)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_ct_uwmed <- df %>%
  filter(!is.na(residuals_ct_uwmed)) %>%
  ggplot(aes(x=residuals_ct_uwmed)) +
  geom_histogram(aes(y = ..density..), binwidth = 1.5) +
  xlab("Residuals") +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$ residuals_ct_uwmed[!is.na(df$residuals_ct_uwmed)])))

gridExtra::grid.arrange(p1_ct_uwmed, p2_ct_uwmed, ncol=2)
```
Again, residuals are not normally distributed, so this glm is not a good model with these primers.


#### Is their a difference in sex by clade?
```{r, echo=FALSE}
table(df$sex, df$clade)
chisq.test(df$sex, df$clade)
```

#### Is there a difference in immunocompromised/cancer by clade?
```{r, echo=FALSE}
table(df$immunocompromised_cancer, df$clade)
chisq.test(df$immunocompromised_cancer, df$clade)
```

#### Is there a difference in inpatient vs. outpatient across clade?
```{r}
table(df$site, df$clade)
chisq.test(df$site, df$clade)
```

#### Is there a difference in proportion of deceased or required ICU care across clade?
```{r}
table(df$severe, df$clade)
chisq.test(df$severe, df$clade)
```

### GLM of inpatient
```{r}
lreg_site = glm(site ~ clade + sex + decade_age + immunocompromised_cancer + week + avg_ct, data = df[!is.na(df$deceased), ], family = "binomial")
summary(lreg_site)

p.adjust(summary(lreg_site)$coefficients[2:7,4], method = "bonferroni", n=6)

df$residuals_site[!is.na(df$bin_age) & !is.na(df$deceased)] = residuals(lreg_site)


p1_site_uwmed <- cbind(Fitted = fitted(lreg_site),Residuals=residuals(lreg_site)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_site_uwmed <- df %>%
  filter(!is.na(residuals_site)) %>%
  ggplot(aes(x=residuals_site)) +
  geom_histogram(aes(y = ..density..), binwidth = 1.5) +
  xlab("Residuals") +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$ residuals_site[!is.na(df$residuals_site)])))

gridExtra::grid.arrange(p1_site_uwmed, p2_site_uwmed, ncol=2)

questionr::odds.ratio(lreg_site, level=(1-(0.05/6)))

or <- as.tibble(questionr::odds.ratio(lreg_site, level=(1-(0.05/6))), rownames = "variable") 

or <- or %>%
  mutate(prob = OR/(OR + 1)) %>%
  rename(lower = "0.4 %") %>%
  rename(upper = "99.6 %") %>%
  mutate(prob2.5 = lower/(lower + 1)) %>%
  mutate(prob97.5 = upper/(upper +1))

or
```
```{r}
colors <- c("#0072B2", "#0072B2", "#0072B2", "#0072B2", "#0072B2", "#0072B2")

plot_or <- or %>%
  filter(variable!="(Intercept)") %>%
  ggplot(aes(x=variable, y=OR)) + 
  geom_errorbar(aes(ymin=lower, ymax=upper), size=0.5, width = 0.2) +
  geom_point(aes(color = variable), size=3) +
  scale_color_manual(values = colors) +
  geom_hline(yintercept=1, linetype="dashed") +
  coord_flip() +
  scale_y_log10() +
  theme_minimal() +
  xlab("") +
  ylab("Odds Ratio for hospitalization") +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("cladeG" = "614G", "decade_age" = "Decade\nincrease in age",  "avg_ct" = "Average Ct", "week" = "Week since start\nof epidemic", "sexM" = "Male", "immunocompromised_canceryes" = "Active cancer or\nimmunocompromised")) +
  geom_text(vjust = 0, aes(label = c("", "", "***", "", "", "*")), size=5)
plot_or
  
```

#### Plotting chart review
```{r}
plotdf<- df %>%
  filter(!is.na(deceased)) %>%
  mutate(female = ifelse(sex == "F", "yes", "no")) %>%
  mutate(inpatient = ifelse(site == "inpatient", "yes", "no")) %>%
  pivot_longer(cols = c(female, immunocompromised_cancer, inpatient, severe), names_to = "variable") %>%
  select(date, clade, variable, value) %>%
  filter(!is.na(variable)) %>%
  mutate(color = ifelse(clade == "D" & value == "yes", "yes.d", NA)) %>%
  mutate(color = replace(color, clade == "D" & value == "no", "no.d")) %>%
  mutate(color = replace(color, clade == "G" & value == "yes", "yes.g")) %>%
  mutate(color = replace(color, clade == "G" & value == "no", "no.g"))

# New facet label names for variable
var.labs <- c("Female", "Immuncompromised \nor active cancer", "Inpatient", "Critical care \n or deceased")
names(var.labs) <- c("female", "immunocompromised_cancer", "inpatient", "severe")

proportions_date <- ggplot(plotdf, aes(x=date)) +
  geom_bar(position = "fill", aes(color=color, fill=color)) +
  geom_vline(xintercept = as.Date("2020-03-09"), linetype=2) +
  geom_vline(xintercept = as.Date("2020-04-02"), linetype=2) +
  scale_fill_manual(values=c('#DCDCDC', '#DCDCDC', '#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#DCDCDC', '#DCDCDC', '#6DC1B3', '#F6C445')) +
  theme_minimal() +
  ylab("Proportion") +
  xlab("Date") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none") +
  facet_grid(clade ~ variable, labeller = labeller(variable = var.labs))
proportions_date
```

Plot proportion by clade
```{r}
stdErr <- function(x, na.rm=FALSE) {
  if (na.rm) x <- na.omit(x)
  sqrt(var(x)/length(x))
}
```

```{r}
plot_clade <- df %>%
  filter(!is.na(deceased) & !is.na(severe)) %>%
  mutate(female = ifelse(sex == "F", "yes", "no")) %>%
  mutate(inpatient = ifelse(site == "inpatient", "yes", "no")) %>%
  pivot_longer(cols = c(female, immunocompromised_cancer, inpatient, severe), names_to = "variable") %>%
  mutate(binary_clade = ifelse(clade == "D", 1, 0)) %>%
  select(clade, variable, value, binary_clade)

plot_clade

stderr <- plot_clade %>%
  dplyr::group_by(variable, value) %>%
  summarise(stderr = stdErr(binary_clade))

errorbars <- plot_clade %>%
  dplyr::group_by(variable, value) %>%
  summarise(mean = mean(binary_clade))%>%
  left_join(stderr) %>%
  mutate(lower = mean - (1.96*stderr)) %>%
  mutate(upper = mean + (1.96*stderr))

errorbars

# New facet label names for variable
var.labs <- c("Female?", "Immunocomp.\nor\nactive cancer?", "Hospitalized?", "Critical care \n or deceased?")
names(var.labs) <- c("female", "immunocompromised_cancer", "inpatient", "severe")

proportion_clade <- plot_clade %>%
  mutate(clade = factor(clade, levels = c("G", "D"))) %>%
  mutate(value = factor(value, levels = c("yes", "no"))) %>%
  ggplot(aes(x=value)) +
  geom_bar(position = "fill", aes(fill = clade, color = clade)) +
  geom_errorbar(data = errorbars, aes(x=value, ymin = lower, ymax = upper), width = 0.5) +
  scale_fill_manual(name = "Clade", labels = c("614G", "614D"), values=c('#F6C445', '#6DC1B3')) +
  scale_color_manual(name = "Clade", labels = c("614G", "614D"), values=c('#F6C445', '#6DC1B3')) +
  theme_minimal() +
  ylab("Proportion") +
  xlab("") +
  theme(legend.position = c(0.72, 1.05),
        legend.direction = "horizontal") +
  facet_wrap(~variable, labeller = labeller(variable = var.labs), strip.position = "bottom", nrow=1) +
  theme(strip.placement = "outside", panel.spacing = unit(0.05, "lines"))
proportion_clade
```
```{r, fig.height=6, fig.width=8.5}
fig_p1 <- ggarrange(dotplot_ct, plot_or, labels = c("A", "C"), nrow=2, ncol=1, font.label = list(size = 20, color = "black", face = "plain", family = NULL), heights=c(1.4,1))
fig_p2 <- ggarrange(table, proportion_clade, labels = c("B", "D"),nrow=2, heights=c(1,1.5), font.label = list(size = 20, color = "black", face = "plain", family = NULL), vjust=c(1.5,0))
figure <- ggarrange(fig_p1, fig_p2, ncol=2, nrow=1, widths=c(1.1, 1))
figure
```

```{r, fig.height=3, fig.width=8}
figure_supplemental <- ggarrange(dotplot_age, age_time, ncol=2, nrow=1, labels=c("A","B"), font.label = list(size = 20, color = "black", face = "plain", family = NULL))
figure_supplemental
```

```{r, fig.height=3, fig.width=8}
figure_supplemental_2 <- ggarrange(orf1_ct, table_orf1, ncol=1, nrow=2, labels=c("A","B"), heights = c(1.5,1), font.label = list(size = 18, color = "black", face = "plain", family = NULL))

annotated <- annotate_figure(figure_supplemental_2,
               top = text_grob("UW Virology: ORF1ab primers", face = "bold", size = 13))

annotated
```
