---
title: "SFS-UW_2020-05-13"
author: "Cassia Wagner"
date: "5/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev="CairoPNG")
options(bitmapType = 'cairo')
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gridExtra)
library(ggsignif)
library(ggpubr)
```
#### Load data:
```{r}
df <- read_tsv("../data/WA_df.tsv")
```
```{r}
summary(df)
```

#### Data prep for analysis:
```{r}
df$primers <- factor(df$primers, levels = c("N1, N2", "ORF1ab", "Egene, RdRp", "wadoh", "scan"))
df$source <- factor(df$source, levels = c("UW", "wadoh", "SFS"))
df$clade <- factor(df$clade, levels = c("D", "G"))
df$county <- factor(df$county)
df$sex <- factor(df$sex)
df$deceased <- factor(df$deceased, levels = c("yes", "no"))
df$site <- factor(df$site, levels = c("outpatient", "inpatient"))
df$critical_care <- factor(df$critical_care)
df$immuncompromised_cancer <- factor(df$immuncompromised_cancer)
df$age_bin = factor(df$age_bin, levels = c("under 10", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80-89", "90+"), ordered = TRUE)
df <- df %>%
  mutate(severe = ifelse(critical_care == "yes" | deceased == "yes", "yes", "no")) %>%
  mutate(severe = factor(severe)) %>%
  mutate(post_symp = date - symptom_onset) %>%
  mutate(ct_norm = avg_ct/avg_rnase_p_crt) %>%
  mutate(king = ifelse(county == "King", "yes", "no")) %>%
  mutate(decade_age = bin_age/10)

df$week[df$date < as.Date("2020-02-25")] = 1
df$week[df$date < as.Date("2020-03-03") & df$date > as.Date("2020-02-24")] = 2
df$week[df$date < as.Date("2020-03-10") & df$date > as.Date("2020-03-02")] = 3
df$week[df$date < as.Date("2020-03-17") & df$date > as.Date("2020-03-09")] = 4
df$week[df$date < as.Date("2020-03-24") & df$date > as.Date("2020-03-16")] = 5
df$week[df$date < as.Date("2020-03-31") & df$date > as.Date("2020-03-23")] = 6
df$week[df$date < as.Date("2020-04-07") & df$date > as.Date("2020-03-30")] = 7
df$week[df$date < as.Date("2020-04-14") & df$date > as.Date("2020-04-06")] = 8
df$week[df$date < as.Date("2020-04-21") & df$date > as.Date("2020-04-13")] = 9
df$week[df$date < as.Date("2020-04-28") & df$date > as.Date("2020-04-20")] = 10
df$week[df$date > as.Date("2020-04-27")] = 12

```
```{r}
summary(df)
```
```{r}
summary(df[df$source == "UW",])
```
```{r}
summary(df[df$source == "SFS",])
```
```{r}
summary(df[df$source == "wadoh",])
```

### CT, AGE, CLADE ANALYSIS
Note: I'm going to split analysis by SFS, UW, WA-DoH sequences because these are two totally sample sets, collected for different reasons and PCR amplifed in different labs. UW sequences are from samples ordered by a healthcare provider while SFS sequences are from people at-home who were not symptomatic enough to seek clinical care.


#### How are the clades spread over time?
```{r, echo=FALSE, fig.height=4, fig.width=8}
labels <- c(UW = "UW\nVirology", wadoh = "WA-DoH", SFS = "SCAN", "(all)" = "All")

clade_time <- df %>%
  ggplot(aes(x=date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.9, 0.8)) +
  ylab("Count") +
  xlab("Date") +
  facet_grid(source ~., margins=TRUE, labeller = labeller(source=labels))
clade_time
```


#### How are UW Virology primers spread over time?
```{r, echo=FALSE, fig.height=4, fig.width=8}
primer_time <- df %>%
  filter(source == "UW") %>%
  ggplot(aes(x=date, color=primers, fill=primers)) + 
  geom_bar(position = position_stack()) + 
  theme_minimal() +
  theme(legend.position = c(0.85, 0.85)) +
  ylab("Count") +
  xlab("Date")
primer_time
```
Do different primers have different Ct's?
```{r, fig.height=4, fig.width=8}
dotplot_ct_primer <- df %>%
  filter(source == "UW") %>%
  ggplot(mapping = aes(x=primers, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, aes(color=primers, fill=primers)) +
  theme_minimal() +
  stat_summary(fun.data="mean_sdl", geom="pointrange", size = 0.1) +
  ylab("Average Ct") +
  xlab("Primer set") +
  geom_signif(comparisons = list(c("Egene, RdRp", "N1, N2"), c("N1, N2", "ORF1ab")), textsize = 3, test = "wilcox.test")
dotplot_ct_primer
```
For UW Virology samples, decided to use N1, N2 primers since they're the most abundant and present when D614 & G614 virus are circulating.

```{r, echo=FALSE, fig.height=4, fig.width=8}
clade_time_n1n2 <- df %>%
  filter(primers == "N1, N2") %>%
  ggplot(aes(x=date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.9, 0.8)) +
  ylab("Count") +
  xlab("Date") +
  labs(subtitle = "Primer set: N1 & N2")
clade_time_n1n2
```



#### Is there a difference in Ct by clade?
For UW:
```{r}
t.test(df$avg_ct[df$source == "UW" & df$primers == "N1, N2"]~df$clade[df$source == "UW" & df$primers == "N1, N2"])
wilcox.test(formula = avg_ct ~ clade, data= df[df$source == "UW" & df$primers == "N1, N2", ], alternative = "two.sided")

median(df$avg_ct[df$source == "UW" & df$primers == "N1, N2" & df$clade == "D"])
median(df$avg_ct[df$source == "UW" & df$primers == "N1, N2" & df$clade == "G"])
```
For SFS:
```{r}
t.test(df$avg_ct[df$source == "SFS"]~df$clade[df$source == "SFS"])
wilcox.test(formula = avg_ct ~ clade, data= df[df$source == "SFS", ], alternative = "two.sided")

median(df$avg_ct[df$source == "SFS" & df$clade == "D"])
median(df$avg_ct[df$source == "SFS" & df$clade == "G"])
```

For WA-DoH:
```{r}
t.test(df$avg_ct[df$source == "wadoh"]~df$clade[df$source == "wadoh"])
wilcox.test(formula = avg_ct ~ clade, data= df[df$source == "wadoh", ], alternative = "two.sided")

median(df$avg_ct[df$source == "wadoh" & df$clade == "D"])
median(df$avg_ct[df$source == "wadoh" & df$clade == "G"])
```

```{r, fig.width=5}
dotplot_ct <- df %>%
  filter(primers == "N1, N2" | source == "SFS" | source == "wadoh") %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, binwidth = 0.5, aes(color=clade, fill=clade)) +
  geom_boxplot(alpha=0, size=0.3) +
  scale_fill_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade", labels = c("614D", "614G"), values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank(),
        panel.spacing = unit(-0.5, "lines")) +
  ylab("Average Ct") +
  xlab("") +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="wilcox.test") +
  facet_wrap(~ source, labeller = labeller(source=labels))
dotplot_ct
```

#### Is there a difference in age by clade?
For UW:
```{r}
t.test(df$bin_age[df$source == 'UW']~df$clade[df$source == 'UW'])
wilcox.test(formula = bin_age ~ clade, data= df[df$source == 'UW', ], alternative = "two.sided")
```
For SFS:
```{r}
t.test(df$bin_age[df$source == 'SFS']~df$clade[df$source == 'SFS'])
wilcox.test(formula = bin_age ~ clade, data= df[df$source == 'SFS', ], alternative = "two.sided")
```

Do we have the power to find an age difference in SCAN samples?
```{r}
testing_age <- function(df, ss) {
  ttest <- df %>%
    sample_n(ss) %>%
    t.test(bin_age ~ clade, data = ., var.equal = TRUE)
  return(ttest$p.value)
}

power_age_79 <- map(1:10000, ~ testing_age(df[df$source == "UW", ], 79))
```
```{r}
sum(power_age_79 < 0.05)/10000
```

```{r, fig.width=5}
dotplot_age <- df %>%
  drop_na(bin_age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=bin_age), na.rm=TRUE) + 
  geom_dotplot(aes( color=clade, fill=clade), binaxis="y", stackdir="center", stackratio = 0.3, dotsize = 0.5, position = position_jitter(width = 0, height=4)) +
  scale_fill_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank()) +
  ylab("Age") +
  xlab("") +
  stat_summary(fun.data="mean_sdl", geom="pointrange", size = 0.1) +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="t.test") +
   facet_wrap(~ source, labeller = labeller(source=labels))
dotplot_age
```



#### Age vs. Ct
```{r, fig.width=7}
ct_age <- df%>%
  filter(source == "UW" & primers == "N1, N2") %>%
  ggplot(aes(x = bin_age, y = avg_ct)) +
  geom_point(aes(color=clade, fill=clade)) +
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  geom_smooth(colour = "black", size=0.5, method = "lm") +
  theme_minimal() +
  xlab("Age") +
  ylab("Average Ct")
ct_age
```

#### Age vs. time
```{r, fig.width=7}
age_time <- df %>%
  drop_na(age_bin) %>%
  ggplot(aes(x = date, y = bin_age)) +
  geom_point(aes(color=clade, fill=clade)) +
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  geom_smooth(method="lm", col = "black", size=0.5) +
  xlab("Date") +
  ylab("Age")
age_time

```


#### Ct vs. time
```{r, fig.width=7}
ct_time <- df %>%
  filter(primers == "N1, N2") %>%
  ggplot(aes(x = date, y = avg_ct)) +
  geom_point(aes(color=clade, fill=clade)) +
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  geom_smooth(method="lm", col = "black", size=0.5) +
  xlab("Date") +
  ylab("Average Ct")
ct_time
```

#### Ct vs. symptom onset
```{r, fig.width=7}
ct_sympon <- df %>%
  filter(!is.na(post_symp)& post_symp <20) %>%
  ggplot(aes(x = post_symp, y = avg_ct)) +
  geom_point(aes(color=clade, fill=clade)) +
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  geom_smooth(data = subset(df, post_symp >=0 & post_symp < 20),method = "lm", color = "black", size=0.5) +
  xlab("Days since symptoms started") +
  ylab("Average Ct")
ct_sympon
```
```{r}
  cor.test(as.numeric(df$post_symp[df$post_symp >= 0 & df$post_symp < 20]), df$avg_ct[df$post_symp >= 0 & df$post_symp < 20], method = "pearson")

  cor.test(as.numeric(df$post_symp[!is.na(df$post_symp)]), df$avg_ct[!is.na(df$post_symp)], method = "pearson")
```


#### Linear model of Ct w/ UW
```{r}
lreg_ct <- glm(avg_ct ~ clade + bin_age + primers + week, data = df[df$source == "UW", ])
summary(lreg_ct)

df$residuals_ct[!is.na(df$age_bin) & df$source == "UW"] = residuals(lreg_ct)

p1_ct <- cbind(Fitted = fitted(lreg_ct), Residuals=residuals(lreg_ct)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_ct <- df %>%
  filter(!is.na(residuals_ct)) %>%
  ggplot(aes(x=residuals_ct)) +
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  xlab("Residuals") +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$ residuals_ct[!is.na(df$residuals_ct)])))

gridExtra::grid.arrange(p1_ct, p2_ct, nrow=1)
```
```{r}
lreg_summary <- tibble("Variable" = c("Intercept", "614G", "Age", "Week", "Egene, RdRp primers", "ORF1ab primers"), 
                       "Coeff. est." = c("20.22", "-1.06", "0.013", "-0.27", "2.51", "-0.29"), 
                       "Std. Error" = c("0.67", "0.26", "0.0056", "0.13", "0.85", "0.35"),
                       "p-value" = c("<2e-16***", "4.6e-05***", "0.025*", "0.035*", "0.0031**", "0.40"))


table <- ggtexttable(lreg_summary, rows = NULL, 
                     theme = ttheme("lBlack", padding = unit(c(4, 3), "mm"), colnames.style = colnames_style(fill = "white", hjust=0, x =0.05, size=10), 
                                    tbody.style = tbody_style(fill = "white", hjust=0, x =0.05, size=10))) 

table
```

```{r}
lreg_ct_n1n2 <- glm(avg_ct ~ clade + bin_age + week, data = df[df$source == "UW" & df$primers == "N1, N2", ])
summary(lreg_ct_n1n2)

df$residuals_ct_n1n2[!is.na(df$age_bin) & df$source == "UW" & df$primers == "N1, N2"] = residuals(lreg_ct_n1n2)

p1_ct_n1n2 <- cbind(Fitted = fitted(lreg_ct_n1n2), Residuals=residuals(lreg_ct_n1n2)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_ct_n1n2 <- df %>%
  filter(!is.na(residuals_ct_n1n2)) %>%
  ggplot(aes(x=residuals_ct_n1n2)) +
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$ residuals_ct[!is.na(df$residuals_ct)])))

gridExtra::grid.arrange(p1_ct_n1n2, p2_ct_n1n2, nrow=1)
```

#### Linear model of Ct with SCAN
```{r}
lreg_ct_sfs <- glm(avg_ct ~ clade + bin_age + week, data = df[df$source == "SFS", ])
summary(lreg_ct_sfs)

df$residuals_ct_sfs[!is.na(df$age_bin) & df$source == "SFS"] = residuals(lreg_ct_sfs)

p1_ct <- cbind(Fitted = fitted(lreg_ct_sfs), Residuals=residuals(lreg_ct_sfs)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_ct <- df %>%
  filter(!is.na(residuals_ct_sfs)) %>%
  ggplot(aes(x=residuals_ct_sfs)) +
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  xlab("Residuals") +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$ residuals_ct_sfs[!is.na(df$residuals_ct_sfs)])))

gridExtra::grid.arrange(p1_ct, p2_ct, nrow=1)
```

### CLINICAL ANALYSIS
```{r}
summary(df[!is.na(df$deceased), ])
```

#### How are the clades spread over time for clinical samples?
```{r, echo=FALSE}
clade_time_uwmed <- df %>%
  filter(!is.na(deceased)) %>%
  ggplot(aes(x=date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  ylab("Count") +
  xlab("Date") +
  labs(subtitle = "Samples from patients in UW Medicine System") +
  facet_grid(primers ~ ., margins = TRUE)
clade_time_uwmed
```

#### Is Ct of UW Med affiliated patients different than Ct of other UW Virology patients?
```{r}
df_uw <- df%>%
  filter(source == "UW") %>%
  mutate(uw = ifelse(!is.na(deceased), "UW affiliated", "Other healthcare network" ))
```
```{r}
df_uw %>%
  filter(primers == "N1, N2") %>%
  {t.test(.$avg_ct ~ .$uw)}
wilcox.test(avg_ct ~ uw, data = df_uw[df_uw$primers == "N1, N2", ])

median(df_uw$avg_ct[df_uw$uw == "Other healthcare network" & df_uw$primers == "N1, N2"])
median(df_uw$avg_ct[df_uw$uw == "UW affiliated" & df_uw$primers == "N1, N2"])
```
```{r}
df_uw %>%
  filter(uw == "Other healthcare network" & primers == "N1, N2") %>%
  wilcox.test(avg_ct ~ clade, data = .)
```
```{r}
dotplot_ct_system <- df_uw %>%
  filter(primers == "N1, N2") %>%
  ggplot(mapping = aes(x=uw, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, binwidth = 0.5, fill="white") +
  geom_boxplot(alpha=0, size=0.3) +
  theme_minimal() +
  ylab("Average Ct") +
  xlab("") +
  geom_signif(comparisons = list(c("Other healthcare network", "UW affiliated")), textsize = 3, test="wilcox.test") 
dotplot_ct_system
```

#### Is Ct different across clade in UW Med affiliated patients?
```{r}
df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  {t.test(.$avg_ct ~ .$clade)}

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
wilcox.test(formula = avg_ct ~ clade, data = ., alternative = "two.sided")

median(df$avg_ct[!is.na(df$deceased) & df$primers == "N1, N2" & df$clade == "D"])
median(df$avg_ct[!is.na(df$deceased) & df$primers == "N1, N2" & df$clade == "G"])
```
```{r}
dotplot_ct_uwmed <- df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.6, binwidth = 0.5, aes(color=clade, fill=clade)) +
  geom_boxplot(alpha=0, size=0.3) +
  scale_fill_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade",values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank()) +
  ylab("Average Ct") +
  xlab("") +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="wilcox.test") +
  labs(subtitle = "Patients at UW Medicine affiliates")
dotplot_ct_uwmed
```


#### Is their a difference in age between UW Medicine patients and other patients tested at UW Virology?
```{r}
t.test(df_uw$bin_age ~ df_uw$uw)
```

#### Is their a difference in age by clade in UW Medicine affiliated patients?
```{r}
df %>%
  filter(!is.na(deceased)) %>%
  {t.test(.$bin_age ~ .$clade)}
```
```{r, fig.width=4, fig.height=4}
dotplot_age_uwmed <- df %>%
  filter(!is.na(deceased)) %>%
  drop_na(bin_age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=bin_age), na.rm=TRUE) + 
  geom_dotplot(aes( color=clade, fill=clade), binaxis="y", stackdir="center", stackratio = 0.8, dotsize = 1, position = position_jitter(width = 0, height=4)) + 
  scale_fill_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  scale_color_manual(name = "Clade", values=c('#6DC1B3', '#F6C445')) +
  theme_minimal() +
  theme(legend.position = c(0.5, -0.045), 
        legend.direction = "horizontal",
        axis.text.x = element_blank()) +
  ylab("Age") +
  xlab("") +
  stat_summary(fun.data="mean_sdl", geom="pointrange", size = 0.1) +
  geom_signif(comparisons = list(c("D", "G")), textsize = 3, test="t.test") +
  labs(subtitle = "Patients at UW Medicine affiliates")
dotplot_age_uwmed
```

#### Across what variables in UW Medicine affiliated patients does Ct differ?
```{r, fig.width=5}
df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = county, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")

df%>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = bin_age, y = avg_ct)) +
  geom_point(aes(color=clade, fill=clade)) +
  geom_smooth(colour = "black", size=0.5, method = "lm")

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = sex, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = immuncompromised_cancer, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = site, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")

df %>%
  filter(!is.na(deceased) & primers == "N1, N2") %>%
  ggplot(aes(x = severe, y = avg_ct)) +
  geom_dotplot(binaxis="y", stackdir="center")
```

#### Linear model of Ct in patients affiliated with UW Medicine
```{r}
lreg_ct_uwmed = glm(avg_ct ~ clade + sex + week + severe + bin_age + site + immuncompromised_cancer, data = df[df$primers == "N1, N2", ])
summary(lreg_ct_uwmed)

df$residuals_ct_uwmed[!is.na(df$bin_age) & !is.na(df$deceased) & df$primers == "N1, N2"] = residuals(lreg_ct_uwmed)

p1_ct_uwmed <- cbind(Fitted = fitted(lreg_ct_uwmed),Residuals=residuals(lreg_ct_uwmed)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_ct_uwmed <- df %>%
  filter(!is.na(residuals_ct_uwmed)) %>%
  ggplot(aes(x=residuals_ct_uwmed)) +
  geom_histogram(aes(y = ..density..), binwidth = 1.5) +
  xlab("Residuals") +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$ residuals_ct_uwmed[!is.na(df$residuals_ct_uwmed)])))

gridExtra::grid.arrange(p1_ct_uwmed, p2_ct_uwmed, ncol=2)
```

#### Is their a difference in sex by clade?
```{r, echo=FALSE}
table(df$sex, df$clade)
chisq.test(df$sex, df$clade)
```

#### Is there a difference in immunocompromised/cancer by clade?
```{r, echo=FALSE}
table(df$immuncompromised_cancer, df$clade)
chisq.test(df$immuncompromised_cancer, df$clade)
```

#### Is there a difference in inpatient vs. outpatient across clade?
```{r}
table(df$site, df$clade)
chisq.test(df$site, df$clade)
```

#### Is there a difference in proportion of deceased or required ICU care across clade?
```{r}
table(df$severe, df$clade)
chisq.test(df$severe, df$clade)
```

### GLM of inpatient
```{r}
lreg_site = glm(site ~ clade + sex + decade_age + immuncompromised_cancer + week + avg_ct, data = df, family = "binomial")
summary(lreg_site)

df$residuals_site[!is.na(df$bin_age) & !is.na(df$deceased)] = residuals(lreg_site)

p1_site_uwmed <- cbind(Fitted = fitted(lreg_site),Residuals=residuals(lreg_site)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

p2_site_uwmed <- df %>%
  filter(!is.na(residuals_site)) %>%
  ggplot(aes(x=residuals_site)) +
  geom_histogram(aes(y = ..density..), binwidth = 1.5) +
  xlab("Residuals") +
  stat_function(fun = dnorm, 
                colour = "red", 
                args = list(mean = 0, sd = sd(df$ residuals_site[!is.na(df$residuals_site)])))

gridExtra::grid.arrange(p1_site_uwmed, p2_site_uwmed, ncol=2)

questionr::odds.ratio(lreg_site)

or <- as.tibble(questionr::odds.ratio(lreg_site), rownames = "variable") 

or <- or %>%
  mutate(prob = OR/(OR + 1)) %>%
  rename(lower = "2.5 %") %>%
  rename(upper = "97.5 %") %>%
  mutate(prob2.5 = lower/(lower + 1)) %>%
  mutate(prob97.5 = upper/(upper +1))

or
```
```{r}
colors <- c("#0072B2", "#0072B2", "#0072B2", "#0072B2", "#0072B2", "#0072B2")

plot_or <- or %>%
  filter(variable!="(Intercept)") %>%
  ggplot(aes(x=variable, y=OR)) + 
  geom_errorbar(aes(ymin=lower, ymax=upper), size=0.5, width = 0.2) +
  geom_point(aes(color = variable), size=3) +
  scale_color_manual(values = colors) +
  geom_hline(yintercept=1, linetype="dashed") +
  coord_flip() +
  scale_y_log10() +
  theme_minimal() +
  xlab("") +
  ylab("Odds Ratio for hospitalization") +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("cladeG" = "614G", "decade_age" = "Decade\nincrease in age",  "avg_ct" = "Average Ct", "week" = "Sample week", "sexM" = "Male", "immuncompromised_canceryes" = "Active cancer or\nimmunocompromised")) +
  geom_text(vjust = 0, aes(label = c("", "", "***", "*", "", "")), size=8)
plot_or
  
```

#### Plotting chart review
```{r}
plotdf<- df %>%
  filter(!is.na(deceased)) %>%
  mutate(female = ifelse(sex == "F", "yes", "no")) %>%
  mutate(inpatient = ifelse(site == "inpatient", "yes", "no")) %>%
  pivot_longer(cols = c(female, immuncompromised_cancer, inpatient, severe), names_to = "variable") %>%
  select(date, clade, variable, value) %>%
  filter(!is.na(variable)) %>%
  mutate(color = ifelse(clade == "D" & value == "yes", "yes.d", NA)) %>%
  mutate(color = replace(color, clade == "D" & value == "no", "no.d")) %>%
  mutate(color = replace(color, clade == "G" & value == "yes", "yes.g")) %>%
  mutate(color = replace(color, clade == "G" & value == "no", "no.g"))

# New facet label names for variable
var.labs <- c("Female", "Immuncompromised \nor active cancer", "Inpatient", "Critical care \n or deceased")
names(var.labs) <- c("female", "immuncompromised_cancer", "inpatient", "severe")

proportions_date <- ggplot(plotdf, aes(x=date)) +
  geom_bar(position = "fill", aes(color=color, fill=color)) +
  geom_vline(xintercept = as.Date("2020-03-09"), linetype=2) +
  geom_vline(xintercept = as.Date("2020-04-02"), linetype=2) +
  scale_fill_manual(values=c('#DCDCDC', '#DCDCDC', '#6DC1B3', '#F6C445')) +
  scale_color_manual(values=c('#DCDCDC', '#DCDCDC', '#6DC1B3', '#F6C445')) +
  theme_minimal() +
  ylab("Proportion") +
  xlab("Date") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none") +
  facet_grid(clade ~ variable, labeller = labeller(variable = var.labs))
proportions_date
```

Plot proportion by clade
```{r}
stdErr <- function(x, na.rm=FALSE) {
  if (na.rm) x <- na.omit(x)
  sqrt(var(x)/length(x))
}
```

```{r}
plot_clade <- df %>%
  filter(!is.na(deceased) & !is.na(severe)) %>%
  mutate(female = ifelse(sex == "F", "yes", "no")) %>%
  mutate(inpatient = ifelse(site == "inpatient", "yes", "no")) %>%
  pivot_longer(cols = c(female, immuncompromised_cancer, inpatient, severe), names_to = "variable") %>%
  mutate(binary_clade = ifelse(clade == "D", 1, 0)) %>%
  select(clade, variable, value, binary_clade)

plot_clade

stderr <- plot_clade %>%
  dplyr::group_by(variable, value) %>%
  summarise(stderr = stdErr(binary_clade))

errorbars <- plot_clade %>%
  dplyr::group_by(variable, value) %>%
  summarise(mean = mean(binary_clade))%>%
  left_join(stderr) %>%
  mutate(lower = mean - (1.96*stderr)) %>%
  mutate(upper = mean + (1.96*stderr))

errorbars

# New facet label names for variable
var.labs <- c("Female?", "Immunocomp.\nor\nactive cancer?", "Hospitalized?", "Critical care \n or deceased?")
names(var.labs) <- c("female", "immuncompromised_cancer", "inpatient", "severe")

proportion_clade <- plot_clade %>%
  mutate(clade = factor(clade, levels = c("G", "D"))) %>%
  mutate(value = factor(value, levels = c("yes", "no"))) %>%
  ggplot(aes(x=value)) +
  geom_bar(position = "fill", aes(fill = clade, color = clade)) +
  geom_errorbar(data = errorbars, aes(x=value, ymin = lower, ymax = upper), width = 0.5) +
  scale_fill_manual(name = "Clade", labels = c("614G", "614D"), values=c('#F6C445', '#6DC1B3')) +
  scale_color_manual(name = "Clade", labels = c("614G", "614D"), values=c('#F6C445', '#6DC1B3')) +
  theme_minimal() +
  ylab("Proportion") +
  xlab("") +
  theme(legend.position = c(0.72, 1.05),
        legend.direction = "horizontal") +
  facet_wrap(~variable, labeller = labeller(variable = var.labs), strip.position = "bottom", nrow=1) +
  theme(strip.placement = "outside", panel.spacing = unit(0.05, "lines"))
proportion_clade
```
```{r, fig.height=6, fig.width=8}
fig_p1 <- ggarrange(dotplot_ct, plot_or, labels = c("A", "C"), nrow=2, ncol=1, font.label = list(size = 20, color = "black", face = "plain", family = NULL), heights=c(1.3,1))
fig_p2 <- ggarrange(table, proportion_clade, labels = c("B", "D"),nrow=2, heights=c(1,1.5), font.label = list(size = 20, color = "black", face = "plain", family = NULL), vjust=c(1.5,0))
figure <- ggarrange(fig_p1, fig_p2, ncol=2, nrow=1, widths=c(1.1, 1))
figure
```

```{r, fig.height=3, fig.width=8}
figure_supplemental <- ggarrange(dotplot_age, age_time, ncol=2, nrow=1, labels=c("A","B"), font.label = list(size = 20, color = "black", face = "plain", family = NULL))
figure_supplemental
```
