---
title: 'ncov-severity: clinical'
author: "Cassia Wagner"
date: "4/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev="CairoPNG")
options(bitmapType = 'cairo')
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

#### Load dataframes
```{r, results='hide'}
df_clade <- read_tsv(file = '../results/df_2020-04-21_UW.tsv')
df_clinical <- read_tsv(file = '../data/20200421_UW_chartreview.tsv')
```

#### Examining df_clade:
- Are any of the samples not in Washington?
```{r}
summarise(df_clade, n())
filter(df_clade, division == 'Washington') %>% summarise(n())
```

- Are collection_date & date the same?
```{r}
df_clade %>% 
  mutate(compare_dates = if_else(collection_date == date, 1, 0)) %>%
  tally()
```

- Are there any missing values?
```{r}
sum(is.na(df_clade))
sum(is.na(df_clade$clade))
df_clade$strain[is.na(df_clade$clade)] 
```

Note: "USA/WA-UW230/2020" is in `ncov/config/exclude.txt` because more tip divergence than expected.
```{r}
df_clade <- drop_na(df_clade)
```

#### When I merge df_clade & df_clinical together, what rows will not merge?
```{r}
anti_join(df_clade, df_clinical, by = c("strain", "collection_date", "age_bin"))
anti_join(df_clinical, df_clade, by = c("strain", "collection_date", "age_bin"))
```

Note: "USA/WA-UW375/2020" and "USA/WA-UW-1697/2020" removed because there was already an identical sequence from each of these patients. 

#### Merging clinical and clade data
```{r}
df <- df_clade %>%
  select(-date, -source, -division) %>%
  inner_join(df_clinical, by = c("strain", "collection_date", "age_bin"))
```

#### Variable prep
```{r, results='hide'}
df$age_bin = factor(df$age_bin, levels = c("11-20", "21-30", "31-40", "41-50", "51-60", "61-70", "71-80", "81-90", "over 90"), ordered = TRUE)
df$clade = factor(df$clade, levels = c("D", "G"))
df$sex = factor(df$sex)
df$deceased = factor(df$deceased, levels = c("yes", "no"))
df$site = factor(df$site)
df$critical_care = factor(df$critical_care, levels = c("yes", "no"))
df$immunocompromised_cancer = factor(df$immunocompromised_cancer, levels = c("yes", "no"))


df$age[df$age_bin == '11-20'] = 15.5
df$age[df$age_bin == '21-30'] = 25.5
df$age[df$age_bin == '31-40'] = 35.5
df$age[df$age_bin == '41-50'] = 45.5
df$age[df$age_bin == '51-60'] = 55.5
df$age[df$age_bin == '61-70'] = 65.5
df$age[df$age_bin == '71-80'] = 75.5
df$age[df$age_bin == '81-90'] = 85.5
df$age[df$age_bin == 'over 90'] = 91 #Better to be conservative
```

#### Data summary
```{r}
summary(df)
```

#### How are the clades spread over time?
```{r, echo=FALSE}
clade_time <- ggplot(df, aes(x=collection_date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Count") +
  xlab("Date")
clade_time
png("../figures/clade-by-date_clinical.png")
clade_time
dev.off()
```

#### Is their a difference in sex by clade?
```{r, echo=FALSE}
table(df$sex, df$clade)
chisq.test(df$sex, df$clade)
```

#### Is there a difference in immunocompromised/cancer by clade?
```{r, echo=FALSE}
table(df$immunocompromised_cancer, df$clade)
chisq.test(df$immunocompromised_cancer, df$clade)
```

#### Is there a difference in age across clade?
```{r}
t.test(df$age ~ df$clade)
wilcox.test(formula = age ~ clade, data= df, alternative = "two.sided")
age <- ggplot(df, mapping = aes(x=clade, y=age_bin, color=clade, fill=clade)) + 
  geom_dotplot(binaxis="y", stackdir="center", binwidth=0.17, position = position_jitter(width = 0, height=0.3)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Samples from patients in UW Medicine system") +
  labs(caption = "D: Mean age = 55.6, G: Mean age = 53.4,  t-test p-value = 0.49\n Wilcoxon Rank Sum test p-value = 0.51")
age
png("../figures/age-clade_clinical.png")
age
dev.off()
```
#### Is there a difference in Ct across clade?
```{r}
t.test(df$avg_ct ~ df$clade)
wilcox.test(formula = avg_ct ~ clade, data= df, alternative = "two.sided")
ct <- ggplot(df, aes(x=clade, y=avg_ct), na.rm=TRUE) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Samples from patients in UW Medicine system") +
  labs(caption = "D: Mean Ct = 18.85, G: Mean Ct = 18.03,  t-test p-value = 0.08\n Wilcoxon Rank Sum test p-value = 0.10")
ct
png("../figures/ct-clade_clinical.png")
ct
dev.off()
```
#### Is there a difference in proportion of deceased across clade?
```{r}
table(df$deceased, df$clade)
chisq.test(df$deceased, df$clade)
```

#### Is there a difference in critical care across clade?
```{r}
table(df$critical_care, df$clade)
fisher.test(df$critical_care, df$clade)
```

#### Is there a difference in inpatient vs. outpatient across clade?
```{r}
table(df$site, df$clade)
chisq.test(df$site, df$clade)
```

Restricting the dataset to the same time frame to the March 10-24 timeframe used for all sequences analyses:
```{r}
df_restricted <- df[df$collection_date > as.Date("2020-03-09") & df$collection_date < as.Date("2020-03-25"), ]
```
```{r}
table(df_restricted$site, df_restricted$clade)
chisq.test(df_restricted$site, df_restricted$clade)
```


### Let severe represent either death or critical care, is there a difference in severe across clade?
```{r}
df <- mutate(df, severe = ifelse(deceased == "yes" | critical_care == "yes", "yes", "no"))

table(df$severe, df$clade)
chisq.test(df$severe, df$clade)
```