---
title: "Prepping WA data D614G analysis"
author: "Cassia Wagner"
date: "December 21, 2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(dev="CairoPNG")
options(bitmapType = 'cairo')
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

#### Load data
```{r}
sfs_df <- read_tsv(file = '../results/df_SFS_2020-08-18.tsv')
uw_df <- read_tsv(file = '../results/df_UW_2020-05-12.tsv')
clinical_df <- read_tsv(file = '../data/20200502_UW_chart_clean.tsv')
early_uw_df <- read_tsv(file = '../results/df_2020-05-02_UW.tsv')
late_uw_df <- read_tsv(file = '../data/df_uw_new_20200811.tsv')
clinical_late <- read_tsv(file = '../data/uw_chartreview_20200811.tsv')
symptoms_df <- read_tsv(file = '../data/sfs_symptoms_2020-05-26.tsv')
wadoh_df <- read_tsv(file='../results/df_wadoh_2020-07-10.tsv')
symp_onset_df <- read_tsv(file='../data/symptom-onset.tsv')
metadata_df <- read_tsv(file='../data/phylodynamics-build-metadata.tsv')
muts_df <- read_tsv(file='../data/subs_and_muts.tsv')
```

#### Data prep

Remove observations outside of Washington
```{r}
early_uw_df <- early_uw_df %>%
  filter(division == "Washington")
uw_df <- uw_df %>%
  filter(division == "Washington")
```

Pre-merge data cleanup
```{r}
early_uw_df <- early_uw_df %>%
  mutate(age = replace(age, age == "unknown", NA))

early_uw_df$age <- as.numeric(early_uw_df$age)

late_uw_df <- late_uw_df %>%
  mutate(source = str_replace(source, "uw", "UW")) %>%
  rename(date = collection_date) %>%
  drop_na(avg_ct)

sfs_df <- sfs_df %>%
  mutate(county = str_remove(location, " County"))

clinical_df <- clinical_df %>%
  select(-X10) %>%
  rename(immunocompromised_cancer = immuncompromised_cancer)

clinical_late <- clinical_late %>%
  select(-X11)

symptoms_df <- symptoms_df %>%
  mutate(sex = replace(sex, sex == "male", "M")) %>%
  mutate(sex = replace(sex, sex == "female", "F"))

wadoh_df <- wadoh_df %>%
  mutate(county = str_remove(location, " County")) %>%
  mutate(county = str_remove(county, " WA"))
```

Are date from Pavitra & date from GISAID the same?
```{r}
uw_df$date_y <- as.Date(uw_df$date_y)
uw_df$date_x <- as.Date(uw_df$date_x)
early_uw_df$date_x <- as.Date(early_uw_df$date_x)
early_uw_df$date_y <- as.Date(early_uw_df$date_y)
```
```{r}
uw_df %>% 
  mutate(compare_dates = if_else(date_x == date_y, 1, 0)) %>%
  filter(is.na(compare_dates) | compare_dates == 0) %>%
  select(strain, date_x, date_y, division, clade, pat_num)
```
```{r}
early_uw_df %>% 
  mutate(compare_dates = if_else(date_x == date_y, 1, 0)) %>%
  filter(is.na(compare_dates) | compare_dates == 0) %>%
  select(strain, date_x, date_y, division, clade, pat_num)
```
Pavitra said to the use date from her. The GISAID date is not correct.
```{r}
early_uw_df <- early_uw_df %>%
  mutate(date = coalesce(date_x, date_y)) %>%
  select(-date_x, -date_y)

uw_df <- uw_df %>%
  mutate(date = coalesce(date_x, date_y)) %>%
  select(-date_x, -date_y)
```


#### Merging together early_uw_df & uw_df

What variables will not merge easily?
```{r}
early_uw_df %>%
  anti_join(uw_df, by = c("strain", "source", "clade", "division", "pat_num", "date"))
uw_df %>%
  anti_join(early_uw_df, by = c("strain", "source", "clade", "division", "pat_num", "date"))
```
```{r}
uw_df %>%
  anti_join(early_uw_df, by = c("strain", "source", "clade", "division", "pat_num", "date", "age_bin"))
```
I will choose to use age from early_uw_df. For "UW366," which I was able to verify using patient number, the earlier date was correct.

```{r}
uw_df %>%
  anti_join(early_uw_df, by = c("strain", "source", "clade", "division", "pat_num", "date", "avg_ct"))
```
avg_ct is essentially all the same; there's just differences due to rounding differences. I'll take Pavitra's averages.


Merging the dataframes
```{r}
df <- early_uw_df %>%
  full_join(uw_df, by = c("strain", "source", "clade", "division", "pat_num", "date")) %>%
  rename(county = client_county) %>%
  mutate(age = coalesce(age.x, age.y)) %>%
  mutate(age_bin = coalesce(age_bin.x, age_bin.y)) %>%
  mutate(avg_ct = coalesce(avg_ct.y, avg_ct.x)) %>%
  mutate(ct_vals = coalesce(ct_vals.y, ct_vals.x)) %>%
  select(-age_bin.x, -age_bin.y, -age.x, -age.y, -avg_ct.y, -avg_ct.x, -ct_vals.y, -ct_vals.x)
```


Adding in primer sets:
```{r}
primers <- str_split(df$ct_vals, "\\|", simplify = TRUE) %>%
  as_tibble() %>%
  mutate(primers = ifelse(V1 != "" & V2 != "","Egene, RdRp", NA)) %>%
  mutate(primers = replace(primers, V6 != "" & V7 != "", "N1, N2")) %>%
  mutate(primers = replace(primers, V5 != "", "ORF1ab")) %>%
  mutate(primers = replace(primers, V5 != "" & V6 != "" & V7 != "", "multiple")) %>%
  mutate(V6 = as.numeric(V6)) %>%
  mutate(V7 = as.numeric(V7))

df <- bind_cols(df, primers)
```
Replacing multiple with "N1, N2" since we'll use "N1, N2" in later analysis.
```{r}
df <- df %>%
  mutate(avg_ct_multiple = (V6+V7)/2) %>%
  mutate(avg_ct = replace(avg_ct, primers == "multiple", avg_ct_multiple[primers == "multiple"])) %>%
  mutate(primers = replace(primers, primers == "multiple", "N1, N2")) %>%
  select(-avg_ct_multiple)
```
```{r}
summary(df$avg_ct)
```

Merging in late UW df:
```{r}
df <- df %>%
  full_join(late_uw_df)
```
Adding in primers variable for late UW data:
```{r}
df <- df %>%
  mutate(primers = replace(primers, EgeneCt != "" & RdRpCt != "","Egene, RdRp")) %>%
  mutate(primers = replace(primers, N1 != "" & N2 != "", "N1, N2")) %>%
  mutate(primers = replace(primers, orf1abCt != "", "ORF1ab")) %>%
  mutate(primers = replace(primers, cobas1 != "" & cobas2 != "", "cobas")) %>%
  mutate(primers = replace(primers, orf1abCt != "" & N1 != "" & N2 != "", "multiple"))
```
Replacing multiple from late_df with "N1, N2" since we'll use "N1, N2" in later analysis.
```{r}
df <- df %>%
  mutate(avg_ct_multiple = (N1+N2)/2) %>%
  mutate(avg_ct = replace(avg_ct, primers == "multiple", avg_ct_multiple[primers == "multiple"])) %>%
  mutate(primers = replace(primers, primers == "multiple", "N1, N2")) %>%
  select(-avg_ct_multiple)
```

#### Merging in clinical data

Merging clinical dataframes together:
```{r}
clinical <- clinical_df %>%
  full_join(clinical_late)
```


Do all samples with patient number have an entry in clinical_df?
```{r}
df %>%
  filter(!is.na(pat_num)) %>%
  anti_join(clinical, by = c("strain", "date"))
```
Note: Both USA/WA-UW375/2020 & USA/WA-UW-1697/2020 are duplicates and the patient already has a record of them. For the other one, the clade is unknown,so it'll be dropped later in the analysis.

When I merge df & clinical_df together, what rows will not merge?
```{r}
anti_join(clinical, df, by = c("strain", "date"))
```


```{r}
df <- df %>%
  left_join(clinical, by = c("strain", "date")) %>%
  mutate(pat_num = coalesce(pat_num.x, pat_num.y)) %>%
  mutate(age = coalesce(age.y, age.x)) %>%
  mutate(age_bin = coalesce(age_bin.y, age_bin.x)) %>%
  select(-pat_num.x, -pat_num.y, -age_bin.x, -age_bin.y, -age.x, -age.y)
```

#### Adding in SFS data
```{r}
df <- df %>%
  bind_rows(sfs_df)
```
```{r}
anti_join(symptoms_df, df, by = "strain")
```
```{r}
df <- df %>%
  left_join(symptoms_df, by = "strain") %>%
  mutate(sex = coalesce(sex.x, sex.y)) %>%
  select(-sex.x, -sex.y)
```
```{r}
df <- df %>%
  full_join(wadoh_df)
```

#### Adding in symptom onset
```{r}
df <- df %>%
  left_join(symp_onset_df, by = "strain") %>%
  mutate(symptom_onset = coalesce(symptom_onset.x, symptom_onset.y)) %>%
  select(-symptom_onset.x, -symptom_onset.y)
```


#### Add in clade
```{r}
df <- df %>%
  rename(spike_614 = clade) 

df <- metadata_df %>%
  select(Strain, Clade) %>%
  rename(clade = Clade, strain = Strain) %>%
  right_join(df, by = "strain")
```

#### Rename UW-379
According to Pavitra, UW379 is no renamed as UW-1328 on GISID, in order for mutations to merge in properly, I need to rename it.
```{r}
df <- df %>%
  mutate(strain = str_replace(strain, "USA/WA-UW379/2020", "USA/WA-UW-1238/2020"))
```


#### Add in mutations
```{r}
muts_df <- muts_df %>%
  rename(strain = Sequence, aa_muts = AA, nt_muts = Mutations) %>%
  select(-Date)

df <- df %>%
  left_join(muts_df, by = "strain")
```


#### Post-merge data cleanup

Check for duplicates:
```{r}
df %>% 
  group_by(strain) %>% 
  filter(n()>1)
```
```{r}
df <- distinct(df)
```
```{r}
df %>% 
  filter(pat_num != "") %>%
  group_by(pat_num) %>% 
  filter(n()>1) %>%
  arrange(pat_num)
```
Will choose to keep the sample that came first.
```{r}
df <- df %>%
  filter(strain != "USA/WA-UW375/2020") %>%
  filter(strain != "USA/WA-UW-2526/2020") %>%
  filter(strain != "USA/WA-UW-4690/2020") %>%
  filter(strain != "USA/WA-UW-1697/2020") %>%
  filter(strain != "USA/WA-UW-822/2020") %>%
  filter(strain != "USA/WA-UW-4931/2020")
```


```{r}
df <- df %>%
  select(-pat_num, -division, -location, -pat_state, -pat_zip, -client_state, -client_zip, -ct_vals, -age, -V1, -V2, -V3, -V4, -V5, -V6, -V7, -EgeneCt, -RdRpCt, -orf1abCt, -cobas1, -cobas2, -N1, -N2, -avg_rnase_p_crt, -clinical_care, -symptoms)
```
```{r}
df <- df %>%
  mutate(symptom_onset = replace(symptom_onset, symptom_onset < as.Date("2020-01-31"), NA)) %>%
  mutate(post_symp = date - symptom_onset) %>%
  mutate(primers = replace(primers, source == "SFS", "scan")) %>%
  mutate(primers = replace(primers, source == "wadoh", "wadoh")) %>%
  mutate(county = replace(county, county == "Unknown", NA))

# Remove symptom onset date
df <- df %>%
  select(-symptom_onset)

# Dropping any samples from May on:
df <- df %>%
  filter(date < as.Date("2020-05-01"))
```


How's it looking?
```{r}
summary(df)
```

For which strains is spike_614 missing?
```{r}
df$strain[is.na(df$spike_614)] 
```
Note: all strains with a missing clade are in ncov/config/exclude.txt, or are filtered out before alignment for other reasons (likely too short a genome?)

```{r}
df <- df %>%
  drop_na(spike_614) %>%
  filter(spike_614 != "other")
```

For which strain is nt_muts missing?
```{r}
df$strain[is.na(df$nt_muts) & df$source == "wadoh"] 
```
These were all removed from GISAID because they were found to be control samples from WA DOH. They should be removed from our dataset too.
```{r}
df$strain[is.na(df$nt_muts) & df$source == "UW"] 
```

```{r}
df <- df %>%
  filter(!(is.na(df$nt_muts) & df$source == "wadoh"))
```

### Make cleaned dataset
```{r}
clean <- df %>%
  select(-post_symp)
```

#### Data overview
```{r}
summary(df)
summary(clean)
```

Saving dataset for analysis:
```{r}
write_tsv(df, "../data/WA_df.tsv")
```
```{r}
write_tsv(clean, "../data/WA_df_cleaned.tsv")
```
