---
title: 'WA-D614G: SFS'
author: "Cassia Wagner"
date: "5/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev="CairoPNG")
options(bitmapType = 'cairo')
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

#### Load data
```{r}
df <- read_tsv(file = '../results/df_SFS_20200513.tsv')
```

## DataQC
How many samples are from Washington?
```{r}
summarise(df, n())
filter(df, division == 'Washington') %>% summarise(n())
```

How many are missing samples?
```{r}
df %>%
  is.na() %>%
    sum()

df %>%
  select(age_bin) %>%
  is.na() %>%
  sum()

df %>%
  select(location) %>%
  is.na() %>%
  sum()

df %>%
  select(clade) %>%
  is.na() %>%
  sum()

df$strain[is.na(df$clade)] 
```

Note: Both "USA/WA-S4/2020" & "USA/WA-S9/2020" were removed in the filtering step.

```{r}
df <- df %>%
  drop_na(clade)
```


Prepping variables
```{r}
df$clade = factor(df$clade)
df$age_bin = factor(df$age_bin, levels = c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80-89"), ordered = TRUE)

df$bin_age[df$age_bin == '0-9'] <- 9 #better to be conservative
df$bin_age[df$age_bin == '10-19'] <- 14.5
df$bin_age[df$age_bin == '20-29'] <- 24.5
df$bin_age[df$age_bin == '30-39'] <- 34.5
df$bin_age[df$age_bin == '40-49'] <- 44.5
df$bin_age[df$age_bin == '50-59'] <- 54.5
df$bin_age[df$age_bin == '60-69'] <- 64.5
df$bin_age[df$age_bin == '70-79'] <- 74.5
df$bin_age[df$age_bin == '80-89'] <- 84.5
```


```{r}
summary(df)
```

```{r, echo=FALSE}
clade_time <- ggplot(df, aes(x=date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Count") +
  xlab("Date")
clade_time
```

#### Is there a difference in Ct by clade?
```{r}
t.test(df$avg_ct ~ df$clade)
wilcox.test(formula = avg_ct ~ clade, data = df, alternative = "two.sided")
```

```{r}
plot_ct <- ggplot(df, aes(x=clade, y=avg_ct)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Sequences from Seattle Flu Study/SCAN") +
  labs(caption = "D: Mean Ct = 25.18,  G: Mean Ct = 23.55,  t-test p-value = 0.114\nWilcoxon Rank Sum test p-value = 0.096")
plot_ct
```
```{r}
dotplot_ct <- df %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 1, binwidth = 1, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom = "crossbar", size = 0.1) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Sequences from Seattle Flu Study/SCAN") +
  labs(caption = "D: Mean Ct = 25.18,  G: Mean Ct = 23.55,  t-test p-value = 0.114\nWilcoxon Rank Sum test p-value = 0.096")
dotplot_ct

```


#### Is their a difference in age by clade?
```{r}
t.test(df$bin_age ~ df$clade)
wilcox.test(formula = bin_age ~ clade, data = df, alternative = "two.sided")
```
```{r}
plot_age_bin <- df %>%
  drop_na(bin_age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=age_bin), na.rm=TRUE) + 
  geom_dotplot(aes( color=clade, fill=clade), binaxis="y", stackdir="center", binwidth= 0.3, dotsize = 2, position = position_jitter(width = 0, height=0)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Sequences from Seattle Flu Study/SCAN") +
  labs(caption = "D: Mean age = 45.4,  G: Mean age = 37.4,  t-test p-value = 0.070\nWilcoxon Rank Sum test p-value = 0.203")
plot_age_bin
```

#### Restricting data to March 8 - April 20
```{r}
df_restricted <- df[df$date > as.Date("2020-03-07") & df$date < as.Date("2020-04-21"), ]
```

#### Is there a difference in Ct by clade for restricted time_frame?
```{r}
t.test(df_restricted$avg_ct ~ df_restricted$clade)
wilcox.test(formula = avg_ct ~ clade, data = df_restricted, alternative = "two.sided")
```

```{r}
plot_ct_restricted <- ggplot(df_restricted, aes(x=clade, y=avg_ct)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Sequences from Seattle Flu Study/SCAN") +
  labs(caption = "D: Mean Ct = 26.00,  G: Mean Ct = 24.09,  t-test p-value = 0.182\nWilcoxon Rank Sum test p-value = 0.172")
plot_ct_restricted
```
```{r}
dotplot_ct_restricted <- df_restricted %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 1, binwidth = 1, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom = "crossbar", size = 0.1) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Sequences from Seattle Flu Study/SCAN") +
  labs(caption = "D: Mean Ct = 26.00,  G: Mean Ct = 24.09,  t-test p-value = 0.182\nWilcoxon Rank Sum test p-value = 0.172")
dotplot_ct_restricted

```

#### Is there a difference in age by clade for restricte timeframe?
#### Is their a difference in age by clade?
```{r}
t.test(df_restricted$bin_age ~ df_restricted$clade)
wilcox.test(formula = bin_age ~ clade, data = df_restricted, alternative = "two.sided")
```

```{r}
# Function for median and IQR

median_IQR <- function(x) {
  data.frame(y = median(x), # Median
             ymin = quantile(x)[2], # 1st quartile
             ymax = quantile(x)[4])  # 3rd quartile
}
```
```{r}
plot_age_bin_restricted <- df_restricted %>%
  drop_na(bin_age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=age_bin), na.rm=TRUE) + 
  geom_dotplot(aes( color=clade, fill=clade), binaxis="y", stackdir="center", binwidth=0.17, dotsize = 2, position = position_jitter(width = 0, height=0.3)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.y = "median") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Sequences from Seattle Flu Study/SCAN") +
  labs(caption = "D: Mean age = 45.6,  G: Mean age = 34.2,  t-test p-value = 0.082\nWilcoxon Rank Sum test p-value = 0.148")
plot_age_bin_restricted
```
