---
title: 'ncov-severity: UW sequences, May 2'
author: "Cassia Wagner"
date: "5/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev="CairoPNG")
options(bitmapType = 'cairo')
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

#### Load dataframes
```{r, results='hide'}
raw <- read_tsv(file = '../results/df_2020-05-02_UW.tsv')
raw_clinical <- read_tsv(file = '../data/20200502_UW_chart_clean.tsv')
```

#### Examining raw:
- Are any of the samples not in Washington?
```{r}
summarise(raw, n())
filter(raw, division == 'Washington') %>% summarise(n())
```
```{r}
raw_wa <- filter(raw, division == 'Washington')
```

- Are date from Pavitra & date from GISAID the same?
```{r}
raw_wa %>% 
  mutate(compare_dates = if_else(date_x == date_y, 1, 0)) %>%
  tally()
```

- Are there any missing values, other than for pat_num?

_Recoding "unknown as na"_
```{r}
raw_wa <- raw_wa %>%
  mutate(age = replace(age, age == "unknown", NA))
```

```{r}
raw_wa %>%
  select(-pat_num) %>%
  is.na() %>%
    sum()

raw_wa %>%
  select(clade) %>%
  is.na() %>%
  sum()

raw_wa$strain[is.na(raw_wa$clade)] 

raw_wa %>%
  select(age) %>%
  is.na() %>%
  sum()

raw_wa %>%
  select(age_bin) %>%
  is.na() %>%
  sum()

raw_wa %>%
  select(date_x) %>%
  is.na() %>%
  sum()
```

Note: all strains with a missing clade are in ncov/config/exclude.txt. Also, I spoke with Pavitra and the for the sequences with a missing date I can use the date from GISAID.
```{r}
raw_wa <- raw_wa %>%
  drop_na(clade)
```
#### Prepping raw df's for merge for merge
```{r}
raw_wa <- raw_wa %>%
  rename(date = date_y)
raw_wa$date <- as.Date(raw_wa$date)
raw_wa$age <- as.numeric(raw_wa$age)
raw_clinical <- raw_clinical %>%
  select(-X10)
```

#### Do all samples with patient number have an entry in raw_clinical?
```{r}
raw_wa %>%
  filter(!is.na(pat_num)) %>%
  anti_join(raw_clinical, by = c("strain", "date"))
```

Note: Both USA/WA-UW375/2020 & USA/WA-UW-1697/2020 are duplicates and the patient already has a record of them.


#### When I merge raw_wa & raw_clinical together, what rows will not merge?
```{r}
anti_join(raw_clinical, raw_wa, by = c("strain", "date"))
```
Note: "USA/WA-UW230/2020" is in `ncov/config/exclude.txt` because more tip divergence than expected.


#### Merging clinical and clade data
```{r}
df <- raw_wa %>%
  select(-date_x, -source, -division, -pat_num) %>%
  left_join(raw_clinical, by = c("strain", "date"))

df <- df %>%
  mutate(age = coalesce(age.y, age.x)) %>%
  mutate(age_bin = coalesce(age_bin.y, age_bin.x)) %>%
  select(-age_bin.x, -age_bin.y, -age.x, -age.y)
```

#### In the full dataframe, how many missing variables do we have?
```{r}
df %>%
  filter(!is.na(deceased)) %>%
  is.na() %>%
  sum()

df %>%
  filter(!is.na(deceased)) %>%
  filter(is.na(critical_care)) %>%
  tally()
```
```{r}
df %>%
  select(-deceased, -site, -critical_care, -sex, -immuncompromised_cancer) %>%
  is.na() %>%
  sum()

df %>%
  filter(is.na(deceased)) %>%
  filter(is.na(age)) %>%
  tally()
  
df %>%
  filter(is.na(deceased)) %>%
  filter(is.na(age_bin)) %>%
  tally()
```

#### Variable prep
```{r, results='hide'}
df$age_bin = factor(df$age_bin, levels = c("under 10", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80-89", "90+"), ordered = TRUE)
df$clade = factor(df$clade, levels = c("D", "G"))
df$sex = factor(df$sex)
df$deceased = factor(df$deceased, levels = c("yes", "no"))
df$site = factor(df$site)
df$critical_care = factor(df$critical_care, levels = c("yes", "no"))
df$immuncompromised_cancer = factor(df$immuncompromised_cancer, levels = c("yes", "no"))

df$bin_age[df$age_bin == 'under 10'] <- 9 #better to be conservative
df$bin_age[df$age_bin == '10-19'] <- 14.5
df$bin_age[df$age_bin == '20-29'] <- 24.5
df$bin_age[df$age_bin == '30-39'] <- 34.5
df$bin_age[df$age_bin == '40-49'] <- 44.5
df$bin_age[df$age_bin == '50-59'] <- 54.5
df$bin_age[df$age_bin == '60-69'] <- 64.5
df$bin_age[df$age_bin == '70-79'] <- 74.5
df$bin_age[df$age_bin == '80-89'] <- 84.5
df$bin_age[df$age_bin == '90+'] <- 91 #Better to be conservative
```

#### Data summary
```{r}
summary(df)
```

## ANALYSIS OF ALL SEQUENCES

#### How are the clades spread over time?
```{r, echo=FALSE}
clade_time <- ggplot(df, aes(x=date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Count") +
  xlab("Date") +
  labs(subtitle = "Sequences from UW Virology")
clade_time
```

#### Is there a difference in Ct by clade?
```{r}
t.test(df$avg_ct~df$clade)
wilcox.test(formula = avg_ct ~ clade, data= df, alternative = "two.sided")
```
```{r}
plot_ct <- ggplot(df, aes(x=clade, y=avg_ct)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Sequences from UW Virology") +
  labs(caption = "D: Mean Ct = 19.81,  G: Mean Ct = 18.23,  t-test p-value = 4e-12\nWilcoxon Rank Sum test p-value = 2e-12")
plot_ct
```




#### Is there a difference in age by clade?
```{r}
t.test(df$age~df$clade)
wilcox.test(formula = age ~ clade, data= df, alternative = "two.sided")
```
```{r}
t.test(df$bin_age~df$clade)
wilcox.test(formula = bin_age ~ clade, data= df, alternative = "two.sided")
```
```{r}
plot_age <- df %>%
  drop_na(age, age_bin) %>%
  ggplot(aes(x=clade, y=age), na.rm=TRUE) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) +
   scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Sequences from UW Virology") +
  labs(caption = "D: Mean age = 57.0,  G: Mean age = 52.7  t-test p-value = 0.003\nWilcoxon Rank Sum test p-value = 0.003")
plot_age
```
```{r}
plot_age_bin <- df %>%
  drop_na(age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=age_bin), na.rm=TRUE) + 
  geom_dotplot(aes( color=clade, fill=clade), binaxis="y", stackdir="center", binwidth=0.17, dotsize = 0.6, position = position_jitter(width = 0, height=0.3)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Sequences from UW Virology") +
  labs(caption = "D: Mean age = 56.9,  G: Mean age = 52.7  t-test p-value = 0.003\nWilcoxon Rank Sum test p-value = 0.004")
plot_age_bin
```


#### What if we restrict analyses by week?
```{r}
df$week[df$date < as.Date("2020-03-10")] = 1 
df$week[df$date < as.Date("2020-03-17") & df$date > as.Date("2020-03-09")] = 2
df$week[df$date < as.Date("2020-03-24") & df$date > as.Date("2020-03-16")] = 3
df$week[df$date < as.Date("2020-03-31") & df$date > as.Date("2020-03-23")] = 4
df$week[df$date < as.Date("2020-04-07") & df$date > as.Date("2020-03-30")] = 5
df$week[df$date > as.Date("2020-04-06")] = 6
week_labels <- c("1" = "Mar 3-9", "2" = "Mar 10-16", "3" = "Mar 17-23", "4" = "Mar 24-30", "5" = "Mar 31 - Apr 6", "6" = "Apr 7-8")
```

```{r}
t.test(df$avg_ct[df$week == 1]~df$clade[df$week == 1])
wilcox.test(formula = avg_ct ~ clade, data= df[df$week == 1, ], alternative = "two.sided")

t.test(df$avg_ct[df$week == 2]~df$clade[df$week == 2])
wilcox.test(formula = avg_ct ~ clade, data= df[df$week == 2, ], alternative = "two.sided")

t.test(df$avg_ct[df$week == 3]~df$clade[df$week == 3])
wilcox.test(formula = avg_ct ~ clade, data= df[df$week == 3, ], alternative = "two.sided")

t.test(df$avg_ct[df$week == 4]~df$clade[df$week == 4])
wilcox.test(formula = avg_ct ~ clade, data= df[df$week == 4, ], alternative = "two.sided")

t.test(df$avg_ct[df$week == 5]~df$clade[df$week == 5])
wilcox.test(formula = avg_ct ~ clade, data= df[df$week == 5, ], alternative = "two.sided")
```

```{r}
ct_by_week <- df %>%
  drop_na(age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  facet_wrap(~ week, labeller=labeller(week = week_labels)) +
  labs(subtitle = "Sequences from UW Virology")
ct_by_week
```
```{r}
dotplot_ct_by_week <- df %>%
  drop_na(age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=avg_ct)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 1.2, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange", size = 0.1) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  facet_wrap(~ week, labeller=labeller(week = week_labels))
dotplot_ct_by_week

```

```{r}
age_by_week <- df %>%
  drop_na(age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=age)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  facet_wrap(~ week, labeller=labeller(week = week_labels))
age_by_week
```

```{r}
age_bin_by_week <- df %>%
  drop_na(age, age_bin) %>%
  ggplot(mapping = aes(x=clade, color=clade, fill=clade)) + 
  geom_dotplot(aes(y=age_bin), binaxis="y", stackdir="center", binwidth=0.1, position = position_jitter(width =0, height = 0.3), dotsize = 2) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  facet_wrap(~ week, labeller=labeller(week = week_labels))
age_bin_by_week
```

#### Restricting data to March 10-27
```{r}
df_restricted <- df[df$date > as.Date("2020-03-09") & df$date < as.Date("2020-03-28"), ]
```
```{r}
summary(df_restricted)
```


#### For data from March 10-27, is there difference in Ct by clade?
```{r}
t.test(df_restricted$avg_ct~df_restricted$clade)
wilcox.test(formula = avg_ct ~ clade, data= df_restricted, alternative = "two.sided")
```
```{r}
plot_ct_restricted <- df_restricted %>%
  drop_na(age, age_bin) %>%
  ggplot(aes(x=clade, y=avg_ct)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Sequences from UW Virology: March 10-27") +
  labs(caption = "D: Mean Ct = 19.56,  G: Mean Ct = 18.33,  t-test p-value = 1e-6\nWilcoxon Rank Sum test p-value = 9e-7")
plot_ct_restricted
```

#### For data from March 10-27, is there a difference in age by clade?
```{r}
t.test(df_restricted$age~df_restricted$clade)
wilcox.test(formula = age ~ clade, data= df_restricted, alternative = "two.sided")
```
```{r}
t.test(df_restricted$bin_age~df_restricted$clade)
wilcox.test(formula = bin_age ~ clade, data= df_restricted, alternative = "two.sided")
```
```{r}
plot_age_restricted <- df_restricted %>%
  drop_na(age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=age)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") + 
  labs(subtitle = "Sequences from UW Virology: March 10-27") +
  labs(caption = "D: Mean age = 56.1,  G: Mean age = 51.9,  t-test p-value = 0.010\nWilcoxon Rank Sum test p-value = 0.013")
plot_age_restricted
```
```{r}
plot_age_bin_restricted <- df_restricted %>%
  drop_na(age, age_bin) %>%
  ggplot(mapping = aes(x=clade, y=age_bin, color=clade, fill=clade)) + 
  geom_dotplot(binaxis="y", stackdir="center", binwidth=0.17, dotsize = 0.85, position = position_jitter(width = 0, height=0.3)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Sequences from UW Virology: March 10-27") +
  labs(caption = "D: Mean age = 55.9,  G: Mean age = 51.9,  t-test p-value = 0.015\nWilcoxon Rank Sum test p-value = 0.017")
plot_age_bin_restricted
```

## CLINICAL ANALYSIS

#### How are the clades spread over time for clinical samples?
```{r, echo=FALSE}
clade_time_clinic <- df %>%
  filter(!is.na(deceased)) %>%
  ggplot(aes(x=date, color=clade, fill=clade)) + 
  geom_bar(position = position_stack()) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Count") +
  xlab("Date") +
  labs(subtitle = "Samples from patients in UW Medicine System")
clade_time_clinic
```



#### Is their a difference in sex by clade?
```{r, echo=FALSE}
table(df$sex, df$clade)
chisq.test(df$sex, df$clade)
```


#### Is there a difference in immunocompromised/cancer by clade?
```{r, echo=FALSE}
table(df$immuncompromised_cancer, df$clade)
chisq.test(df$immuncompromised_cancer, df$clade)
```

#### Is there a difference in Ct across clade?
```{r}
t.test(df$avg_ct[!is.na(df$deceased)] ~ df$clade[!is.na(df$deceased)])
wilcox.test(formula = avg_ct ~ clade, data = df[!is.na(df$deceased),], alternative = "two.sided")
```
```{r}
plot_ct_clinic <- df %>%
  filter(!is.na(deceased)) %>%
ggplot(aes(x=clade, y=avg_ct), na.rm=TRUE) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Samples from patients in UW Medicine system") +
  labs(caption = "D: Mean Ct = 18.83,  G: Mean Ct = 17.99,  t-test p-value = 0.066\nWilcoxon Rank Sum test p-value = 0.056")
plot_ct_clinic
```

#### Is there a difference in age across clade?
```{r}
t.test(df$age[!is.na(df$deceased)] ~ df$clade[!is.na(df$deceased)])
wilcox.test(formula = age ~ clade, data= df[!is.na(df$deceased),], alternative = "two.sided")
```
```{r}
t.test(df$bin_age[!is.na(df$deceased)] ~ df$clade[!is.na(df$deceased)])
wilcox.test(formula = bin_age ~ clade, data= df[!is.na(df$deceased),], alternative = "two.sided")
```
```{r}
plot_age_clinic <- df %>%
  filter(!is.na(deceased)) %>%
  ggplot(mapping = aes(x=clade, y=age)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Samples from patients in UW Medicine system") +
  labs(caption = "D: Mean age = 55.7,  G: Mean age = 52.4,  t-test p-value = 0.281\nWilcoxon Rank Sum test p-value = 0.309")
plot_age_clinic
```
```{r}
plot_age_bin_clinic <- df %>%
  filter(!is.na(deceased)) %>%
  ggplot(mapping = aes(x=clade, y=age_bin, color=clade, fill=clade)) + 
  geom_dotplot(binaxis="y", stackdir="center", binwidth=0.17, position = position_jitter(width = 0, height=0.3)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Samples from patients in UW Medicine system")  +
  labs(caption = "D: Mean age = 55.5,  G: Mean age = 52.5,  t-test p-value = 0.321\nWilcoxon Rank Sum test p-value = 0.332")
plot_age_bin_clinic
```

#### Is there a difference in proportion of deceased across clade?
```{r}
table(df$deceased, df$clade)
chisq.test(df$deceased, df$clade)
```

#### Is there a difference in critical care across clade?
```{r}
table(df$critical_care, df$clade)
fisher.test(df$critical_care, df$clade)
```

#### Is there a difference in inpatient vs. outpatient across clade?
```{r}
table(df$site, df$clade)
chisq.test(df$site, df$clade)
```

### Let severe represent either death or critical care, is there a difference in severe across clade?
```{r}
df <- mutate(df, severe = ifelse(deceased == "yes" | critical_care == "yes", "yes", "no"))

table(df$severe, df$clade)
chisq.test(df$severe, df$clade)
```


### Restricting clinical data to March 10-27

#### Is their a difference in sex by clade?
```{r, echo=FALSE}
table(df_restricted$sex, df_restricted$clade)
chisq.test(df_restricted$sex, df_restricted$clade)
```

#### Is there a difference in immunocompromised/cancer by clade?
```{r, echo=FALSE}
table(df_restricted$immuncompromised_cancer, df_restricted$clade)
chisq.test(df_restricted$immuncompromised_cancer, df_restricted$clade)
```


#### Is there a difference in Ct across clade?
```{r}
t.test(df_restricted$avg_ct[!is.na(df_restricted$deceased)] ~ df_restricted$clade[!is.na(df_restricted$deceased)])
wilcox.test(formula = avg_ct ~ clade, data = df_restricted[!is.na(df_restricted$deceased),], alternative = "two.sided")
```
```{r}
plot_ct_clinic_restricted <- df_restricted %>%
  filter(!is.na(deceased)) %>%
ggplot(aes(x=clade, y=avg_ct), na.rm=TRUE) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Samples from patients in UW Medicine system: March 10-27") +
  labs(caption = "D: Mean Ct = 18.68,  G: Mean Ct = 17.88,  t-test p-value = 0.126\nWilcoxon Rank Sum test p-value = 0.163")
plot_ct_clinic_restricted
```
```{r}
dotplot_ct <- df_restricted %>%
  filter(!is.na(deceased)) %>%
  ggplot(mapping = aes(x=clade, y=avg_ct, color=clade, fill=clade)) + 
  geom_dotplot(binaxis="y", stackdir="center", dotsize = 0.5, binwidth = 0.8) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Average Ct") +
  xlab("Clade") +
  labs(subtitle = "Samples from patients in UW Medicine system: March 10-27") +
  labs(caption = "D: Mean Ct = 18.68,  G: Mean Ct = 17.88,  t-test p-value = 0.126\nWilcoxon Rank Sum test p-value = 0.163")
dotplot_ct
```

#### Is there a difference in age across clade?
```{r}
t.test(df_restricted$age[!is.na(df_restricted$deceased)] ~ df_restricted$clade[!is.na(df_restricted$deceased)])
wilcox.test(formula = age ~ clade, data= df_restricted[!is.na(df_restricted$deceased),], alternative = "two.sided")
```
```{r}
t.test(df_restricted$bin_age[!is.na(df_restricted$deceased)] ~ df_restricted$clade[!is.na(df_restricted$deceased)])
wilcox.test(formula = bin_age ~ clade, data= df_restricted[!is.na(df_restricted$deceased),], alternative = "two.sided")
```
```{r}
plot_age_clinic_restricted <- df_restricted %>%
  filter(!is.na(deceased)) %>%
  ggplot(mapping = aes(x=clade, y=age)) + 
  geom_violin(alpha=0.7, aes(color=clade, fill=clade)) + 
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  stat_summary(fun.data="mean_sdl", geom="pointrange") +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Samples from patients in UW Medicine system: March 10-27") +
  labs(caption = "D: Mean age = 56.2,  G: Mean age = 51.7,  t-test p-value = 0.208\nWilcoxon Rank Sum test p-value = 0.228")
plot_age_clinic_restricted
```
```{r}
plot_age_bin_clinic_restricted <- df_restricted %>%
  filter(!is.na(deceased)) %>%
  ggplot(mapping = aes(x=clade, y=age_bin, color=clade, fill=clade)) + 
  geom_dotplot(binaxis="y", stackdir="center", binwidth=0.17, position = position_jitter(width = 0, height=0.3)) + 
  scale_color_manual(values=c('#4C90C0', '#AEBD50')) +
  scale_fill_manual(values=c('#4C90C0', '#AEBD50')) +
  theme_minimal() +
  theme(text = element_text(family = "NimbusSan", size = 10)) +
  ylab("Age") +
  xlab("Clade") +
  labs(subtitle = "Samples from patients in UW Medicine system: March 10-27") +
  labs(caption = "D: Mean age = 55.9,  G: Mean age = 51.9,  t-test p-value = 0.262\nWilcoxon Rank Sum test p-value = 0.278")
plot_age_bin_clinic_restricted
```


#### Is there a difference in proportion of deceased across clade?
```{r}
table(df_restricted$deceased, df_restricted$clade)
chisq.test(df_restricted$deceased, df_restricted$clade)
```

#### Is there a difference in critical care across clade?
```{r}
table(df_restricted$critical_care, df_restricted$clade)
fisher.test(df_restricted$critical_care, df_restricted$clade)
```

### Is there a difference in inpatient vs. outpatient across clade?
```{r}
table(df_restricted$site, df_restricted$clade)
chisq.test(df_restricted$site, df_restricted$clade)
```

### Let severe represent either death or critical care, is there a difference in severe across clade?
```{r}
df_restricted <- mutate(df_restricted, severe = ifelse(deceased == "yes" | critical_care == "yes", "yes", "no"))


table(df_restricted$severe, df_restricted$clade)
chisq.test(df_restricted$severe, df_restricted$clade)
```
